---
title: "Data preparation"
output: html_document
date: "2025-01-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("D:/Users/e1002963/Local/NORSE")

library(readxl)
library(reshape2)
library(plyr)
library(ggplot2)
library(terra)
library(sf)

```

# Epi- and infauna

Reordering all the data and combining it into long format. Keeping only variables needed for analysis.


## Finland

Notes on data editing

Epifauna
- Counts and counts per m2. All counts transformed to m2
- Also for those all Zostera DW and shoot densities transformed to m2
- Floats rounded as earlier were as well
- One abundance marked as x, changed to NA
- Remove observations with no Zostera

Infauna
- Sieve assumed based on comment

Tvärminne
- Assuming old data is already per m2
- Using 1971 as year for old data

```{r}
fin <- as.data.frame(read_xlsx("data/EelgrassFaunaData-Finland.xlsx",
                               sheet = "EpifaunaData"))
var_fin <- c("Country", "Site", "Year", "Month",
             "Reference ID (from \"DataSources\" sheet)",
             "Number of replicates (N)", "Raw data or means?",
             "Replicate # (if raw data) from 1 to N", "Epifauna or Infauna",
             "Sampling method", "Mesh/sieve size (mm)", "Sample size (m-2)",
             "Zostera dw", "Zostera shoot density (m-2)", "Notes")

fin_l <- melt(fin, id.vars = var_fin, variable.name = "Taxa",
              value.name = "Abundance", factorsAsStrings = TRUE)

fin_l$Abundance <- round(as.numeric(fin_l$Abundance))
fin_l[is.na(fin_l$Abundance), "Abundance"] <- 0

## Standardising abundance to M2
fin_l <- fin_l[!fin_l$Notes %in% "REMOVE (no Zostera)", ]
fin_l[!fin_l$Notes %in% "Data already standardised to m-2", ]$Abundance <-
  1 / fin_l[!fin_l$Notes %in% "Data already standardised to m-2", ]$`Sample size (m-2)` *
  fin_l[!fin_l$Notes %in% "Data already standardised to m-2", ]$Abundance

## Standardising Zostera DW to M2
fin_l$`Zostera dw (m-2)` <- fin_l$`Zostera dw`
fin_l[!fin_l$Notes %in% "Data already standardised to m-2", ]$`Zostera dw (m-2)` <-
  1 / fin_l[!fin_l$Notes %in% "Data already standardised to m-2", ]$`Sample size (m-2)` *
  fin_l[!fin_l$Notes %in% "Data already standardised to m-2", ]$`Zostera dw (m-2)`

## Standardising Zostera shoot count to M2
fin_l[!fin_l$Notes %in% "Data already standardised to m-2", ]$`Zostera shoot density (m-2)` <-
  1 / fin_l[!fin_l$Notes %in% "Data already standardised to m-2", ]$`Sample size (m-2)` *
  fin_l[!fin_l$Notes %in% "Data already standardised to m-2", ]$`Zostera shoot density (m-2)`

fin_l$Taxa <- as.character(fin_l$Taxa)

### Infauna
fin_in <- as.data.frame(read_xlsx("data/EelgrassFaunaData-Finland.xlsx",
                                  sheet = "InfaunaData"))

names(fin_in)[!names(fin_in) %in% names(fin_l)]
var_fin_in <- c(var_fin[c(var_fin) %in% names(fin_in)], "Sieve size (mm)")

fin_lin <- melt(fin_in, id.vars = var_fin_in, variable.name = "Taxa",
                value.name = "Abundance", factorsAsStrings = TRUE)

fin_lin$Abundance <- as.numeric(fin_lin$Abundance)
fin_lin[is.na(fin_lin$Abundance), "Abundance"] <- 0

## Edits
fin_lin[fin_lin$Notes %in% "sieve size is probably 0.5, as that is the standard sieve size for infauna in finalnd", ]$`Sieve size (mm)` <- 0.5
fin_lin$`Sieve size (mm)` <- as.numeric(gsub(",", ".",
                                             fin_lin$`Sieve size (mm)`))

## Abundance
fin_lin[fin_lin$`Sample size (m-2)` == "1 (0.00173)", "Sample size (m-2)"] <- 0.00173
fin_lin$`Sample size (m-2)` <- as.numeric(gsub(",", ".",
                                               fin_lin$`Sample size (m-2)`))

fin_lin[!fin_lin$Notes %in%
          "Data already standardised to m-2", ]$Abundance <-
  1 /
  fin_lin[!fin_lin$Notes %in% "Data already standardised to m-2", ]$`Sample size (m-2)` *
  fin_lin[!fin_lin$Notes %in% "Data already standardised to m-2", ]$Abundance

fin_lin$Taxa <- as.character(fin_lin$Taxa)

########################### BOTH
### Infauna
fin_3 <- as.data.frame(read_xlsx("data/EelgrassFaunaData-Finland.xlsx",
                                 sheet = "TvärminneSamplerData", n_max = 10))


fin_3[!fin_3$`Raw data or means?` %in% "mean", ]$`Raw data or means?` <- "Raw data"

names(fin_3)[!names(fin_3) %in% names(fin_l)]
var_fin_in <- c(var_fin[c(var_fin) %in% names(fin_3)], "Mesh size (mm)")

fin_3 <- melt(fin_3, id.vars = var_fin_in, variable.name = "Taxa",
              value.name = "Abundance", factorsAsStrings = TRUE)

fin_3$Abundance <- as.numeric(fin_3$Abundance)
fin_3[is.na(fin_3$Abundance), "Abundance"] <- 0

## Edits
fin_3$Year <- as.numeric(gsub("1968-71", "1971", fin_3$Year))

## Abundance
fin_3[fin_3$`Raw data or means?` %in% "Raw data", "Abundance"] <-   1 /
  fin_3[fin_3$`Raw data or means?` %in% "Raw data", ]$`Sample size (m-2)` *
  fin_3[fin_3$`Raw data or means?` %in% "Raw data", ]$Abundance

# fin_3[fin_3$`Raw data or means?` %in% "mean", "Zostera dw"] <-   1 /
#   fin_3[fin_3$`Raw data or means?` %in% "mean", ]$`Sample size (m-2)` *
#   fin_3[fin_3$`Raw data or means?` %in% "mean", ]$`Zostera dw`

fin_3$Taxa <- as.character(fin_3$Taxa)
fin_3$`Epifauna or Infauna` <- "Combined"

###############



fin_both <- merge(fin_l, fin_lin, all = TRUE)
fin_both <- merge(fin_both, fin_3, all = TRUE)


fin_both$Country <- "Finland"
fin_both[fin_both$Site %in% c("Södra Stenskäret, Askö"), "Country"] <- "Sweden E"

fin_both[fin_both$Site %in% c("Södra Stenskäret, Askö"), "Site"] <- "Stenskäret"
fin_both[fin_both$`Sampling method` %in% "mesh bag", "Sampling method"] <- "Mesh bag"


```




## Sweden
Notes on data editing
- Assuming abundance is per sample
- xs is foramnifera
- Country set as east coast vs. west coast


```{r}
swe <- as.data.frame(read_xlsx("data/EelgrassFauna-Sweden.xlsx", sheet = "Epifauna", skip = 1))
var_swe <- c("Country", "Site", "Year", "Month",
             "Reference ID Number", "Sample size (m-2)",
             "Reference ID (from \"DataSources\" sheet)",
             "Number of replicates (N)", "Raw data or means?",
             "Replicate # (if raw data) from 1 to N", "Epifauna or Infauna",
             "Sampling method", "Sieve size (mm)", "Zostera dw, g/sample",
             "Zostera shoot density", "Epiphytes", "Filamentous algae (not epiphytes)",
             "Fucus serratus dw(g)/sample", "Fucus vesiculosus dw/sample",
             "Cladophora/Chlorella dw/sample", "Notes", "Depth", "Potamogeton pectinatus",
             "Zostera dw (m-2)", "Zostera shoot density (m-2)")

names(swe)[names(swe) %in% "Mesh bag"] <- "Sampling method"
swe[swe$`Sample size (m-2)` %in% "35cm x 35 cm =0,1225 m2 (x med 8,16 för 1 m2)", "Sample size (m-2)"] <- 0.1225
swe$`Sample size (m-2)` <- as.numeric(swe$`Sample size (m-2)`)

swe <- swe[!swe$`Replicate # (if raw data) from 1 to N` %in% "SD", ]
swe$`Zostera shoot density (m-2)` <- gsub(" ", "", swe$`Zostera shoot density`)
swe$`Zostera shoot density (m-2)` <- gsub(",", ".", swe$`Zostera shoot density (m-2)`)
swe$`Zostera shoot density (m-2)` <- as.numeric(gsub("/m2", "", swe$`Zostera shoot density (m-2)`))

swe$`Zostera dw (m-2)` <-  1 / swe$`Sample size (m-2)` * as.numeric(swe$`Zostera dw, g/sample`)
swe$`Zostera dw, g/sample` <- gsub(",", ".", swe$`Zostera dw, g/sample`)
swe$`Zostera dw, g/sample` <- as.numeric(gsub(" gDW/m2", "", swe$`Zostera dw, g/sample`))
swe[is.na(swe$`Zostera dw (m-2)`), "Zostera dw (m-2)"] <- swe[is.na(swe$`Zostera dw (m-2)`), "Zostera dw, g/sample"]

swe$xs <- NULL

swe_l <- melt(swe, id.vars = var_swe, variable.name = "Taxa", value.name = "Abundance",
              factorsAsStrings = TRUE)

swe_l$Abundance <- as.numeric(swe_l$Abundance)
swe_l$Abundance <- 1 / swe_l$`Sample size (m-2)` * swe_l$Abundance
swe_l$Taxa <- as.character(swe_l$Taxa)
swe_l[swe_l$`Sample size (m-2)` %in% 1, "Sample size (m-2)"] <- 0.1225


swe_l$Country <- "Sweden W"
swe_l[swe_l$Site %in% c("Sodra Stenskaret"), "Country"] <- "Sweden E"

swe_l$Taxa <- gsub("_", " ", swe_l$Taxa)

swe_l[swe_l$Site %in% c("Finsbo (to be analysed)"), "Site"] <- "Finsbo"
swe_l[swe_l$Site %in% c("Lindholmen (to be analysed)"), "Site"] <- "Lindholmen"
swe_l[swe_l$Site %in% c("Lindholmen, Fucus i Zost."), "Site"] <- "Lindholmen"
swe_l[swe_l$Site %in% c("Sodra Stenskaret"), "Site"] <- "Stenskäret"


## Infauna
swe_in <- as.data.frame(read_xlsx("data/EelgrassFauna-Sweden.xlsx",
                                  sheet = "Infauna"))

names(swe_in)[!names(swe_in) %in% names(swe_l)]
var_swe_in <- c(var_swe[c(var_swe) %in% names(swe_in)])
swe_in <- swe_in[, !names(swe_in) %in% c("Zostera dw",
                                         "Etc....104", "Etc....105",
                                         "Etc....106", "Etc....107",
                                         "Etc....108", "Etc....109",
                                         "Notes...110", "Notes...111")]


swe_lin <- melt(swe_in, id.vars = var_swe_in,
                variable.name = "Taxa",
                value.name = "Abundance",
                factorsAsStrings = TRUE)

## 10.4 cm diameter and 8 cm depth
circle_area <- pi * (0.104 / 2)^2 
m2_conversion <- 1 / circle_area

swe_lin$Abundance <- as.numeric(swe_lin$Abundance)
swe_lin[swe_lin$`Sample size (m-2)` %in% c("x 105,2 to per m2",
                                           "x 105,2 to per m2, 95,02 cm2, 11 cm diam"), "Abundance"] <-
  swe_lin[swe_lin$`Sample size (m-2)` %in% c("x 105,2 to per m2",
                                           "x 105,2 to per m2, 95,02 cm2, 11 cm diam"), "Abundance"] *
  105.2

swe_lin[swe_lin$Site %in% c("Finsbo  (obs means per m2)"), "Abundance"] <-
  swe_lin[swe_lin$Site %in% c("Finsbo  (obs means per m2)"), "Abundance"] * m2_conversion

## Correct the m2
swe_lin[swe_lin$`Sample size (m-2)` %in% c("x 105,2 to per m2",
                                           "x 105,2 to per m2, 95,02 cm2, 11 cm diam"),
        "Sample size (m-2)"] <- 0.0095033

swe_lin[swe_lin$Site %in% c("Finsbo  (obs means per m2)"), "Sample size (m-2)"] <- circle_area


swe_lin[swe_lin$Site %in% c("Finsbo  (obs means per m2)"), "Site"] <- "Finsbo"
swe_lin$Taxa <- as.character(swe_lin$Taxa)

swe_lin$Country <- "Sweden W"

swe_both <- merge(swe_l, swe_lin, all = TRUE)

```

## Germany

Notes on data editing
- Assuming raw counts
- Negative abundances recoded as 0
- Epifauna

```{r}

de <- as.data.frame(read_xlsx("data/EelgrassFaunaData-Germany.xlsx", sheet = "Epifauna"))
var_de <- c("STATION_NR", "Site", "ORT", "HOL_CODE", "HOL_DEPTH", "DAY", "YEAR",
            "MONTH", "Sampling method", "rw_geo_mess", "REMARK", "hw_geo_mess",
            "Sample size (m-2)", "Mesh Size (mm)")

names(de)[names(de) %in% "Sample Size (m2)"] <- "Sample size (m-2)"
names(de)[names(de) %in% "SAMPLING DEVICE"] <- "Sampling method"

de_l <- melt(de, id.vars = var_de, variable.name = "Taxa", value.name = "Abundance",
             factorsAsStrings = TRUE)

de_l[de_l$Abundance < 0, "Abundance"] <- 0
de_l$Abundance <- 1 / de_l$`Sample size (m-2)` * de_l$Abundance
de_l$Taxa <- as.character(de_l$Taxa)

de_l$`Epifauna or Infauna` <- "Epifauna"

```


## Norway
Notes on data editing
- Dropping observations with values as cr and x
- Editing 1996 data as replicates

```{r}
nor <- as.data.frame(read_xlsx("data/EelgrassFaunaData-Norway.xlsx", sheet = "Epifauna",
                               skip = 1))
nor <- nor[, !names(nor) %in% c("...234", "...235", "...236",
                                "...237", "...238", "...239",
                                "...240", "...241", "...242",
                                "...243", "...244", "...245",
                                "...246", "...247", "...248",
                                "...249", "...250", "...251")]

var_nor <- c("Country", "Site", "Year", "Month",
             "Reference ID (from \"DataSources\" sheet)",
             "Number of replicates (N)", "Raw data or means?",
             "Replicate # (if raw data) from 1 to N", "Epifauna or Infauna",
             "Mesh bag", "Sieve size (mm)", "Sample size (m-2)",
             "Zostera dw, g/sample", "Zostera shoot density", "Notes")

nor[nor$`Sample size (m-2)` %in% "Density (abundace per 1/4 m2 quadrat)", "Sample size (m-2)"] <- 0.25
nor$`Sample size (m-2)` <- as.numeric(nor$`Sample size (m-2)`)

nor_l <- melt(nor, id.vars = var_nor, variable.name = "Taxa", value.name = "Abundance",
              factorsAsStrings = TRUE)

nor_l <- nor_l[!nor_l$Abundance %in% c("cr", "x"), ]
nor_l$Abundance <- as.numeric(nor_l$Abundance)
nor_l$Abundance <- 1 / nor_l$`Sample size (m-2)` * nor_l$Abundance

## Correct the sample size based on Lauras comment in file
nor_l[nor_l$Year %in% 2023, "Sample size (m-2)"] <- 0.355

nor_l$Taxa <- as.character(nor_l$Taxa)



nor_l[nor_l$Site == "Arendal (station Klaua/Klauva) (Station 1)",
      "Replicate # (if raw data) from 1 to N"] <- 1
nor_l[nor_l$Site == "Arendal (station Klaua/Klauva) (Station 2)",
      "Replicate # (if raw data) from 1 to N"] <- 2
nor_l[nor_l$Site == "Arendal (station Klaua/Klauva) (Station 3)",
      "Replicate # (if raw data) from 1 to N"] <- 3

## Renaming sites to match cooridinate file
nor_l[nor_l$Site %in% c("Klava",
                        "Arendal (station Klaua/Klauva) (Station 1)",
                        "Arendal (station Klaua/Klauva) (Station 2)",
                        "Arendal (station Klaua/Klauva) (Station 3)"), "Site"] <- "Klauva"

nor_l[nor_l$Site %in% c("Station 146, Risør ytre fjord"),
      "Site"] <- "Risør ytre fjord"
nor_l[nor_l$Site %in% c("Station 122, Barmen (Risør midtre fjord)"),
      "Site"] <- "Risør midtre fjord (Barmen)"
nor_l[nor_l$Site %in% c("Station 111, Sørfjorden (Risør indre fjord)"),
      "Site"] <- "Risør indre fjord (Sørfjorden)"

nor_l[nor_l$Site %in% c("Station 200, Langesund"),
      "Site"] <- "Langesund"
# nor_l[nor_l$Site %in% c("Station 235, Nøtterøy"),
#       "Site"] <- ""
nor_l[nor_l$Site %in% c("Station 192, Eidanger"),
      "Site"] <- "Eidanger"
nor_l[, "Sampling method"] <- "Mesh bag"

## Infauna
nor_in <- as.data.frame(read_xlsx("data/EelgrassFaunaData-Norway.xlsx",
                                  sheet = "Infauna"))

names(nor_in)[!names(nor_in) %in% names(nor)]

nor_in$`Zostera dw` <- NULL
nor_in$`Sampling method` <- "Core"
var_nor_in <- c(var_nor[c(var_nor) %in% names(nor_in)], "Sampling method")

nor_lin <- melt(nor_in, id.vars = var_nor_in,
                variable.name = "Taxa",
                value.name = "Abundance",
                factorsAsStrings = TRUE)

nor_lin$Taxa <- as.character(nor_lin$Taxa)

## Combine
nor_both <- merge(nor_l, nor_lin, all = TRUE)


```


## Estonia

Notes on data editing
- No raw data and not possible to backtransform
- Zostera dw and shoot also per m2?

```{r}
est <- as.data.frame(read_xlsx("data/NORSE-WP2EelgrassFaunaTemplate-Estonia241010.xlsx",
                               sheet = "Data"))

var_est <- c("Country", "Site", "Year", "Month",
             "Sample size (m-2)",
             "Reference ID (from \"DataSources\" sheet)",
             "Number of replicates (N)", "Raw data or means?",
             "Replicate # (if raw data) from 1 to N", "Epifauna or Infauna",
             "Sampling method", "Sieve size (mm)", "Zostera dw",
             "Zostera shoot density", "Depth", "Comments")

names(de)[names(de) %in% "Zostera dw"] <- "Zostera dw (m-2)"
names(de)[names(de) %in% "Zostera shoot density"] <- "Zostera shoot density (m-2)"

est_l <- melt(est, id.vars = var_est, variable.name = "Taxa", value.name = "Abundance",
              factorsAsStrings = TRUE)
est_l$Taxa <- as.character(est_l$Taxa)
est_l$`Sample size (m-2)` <- 0.04

```

## Combine data

```{r}
# Combining the columns needed for analysis
## All var from fin

names(fin_both)[!names(fin_both) %in% names(swe_both)]

## Renaming the sieve size variable 
names(swe_both)[names(swe_both) == "Sieve size (mm)"] <- "Mesh/sieve size (mm)"
names(de_l)[names(de_l) == "Mesh Size (mm)"] <- "Mesh/sieve size (mm)"
names(nor_both)[names(nor_both) == "Sieve size (mm)"] <- "Mesh/sieve size (mm)"
names(est_l)[names(est_l) == "Sieve size (mm)"] <- "Mesh/sieve size (mm)"


## Renaming de data
de_l$Country <- "Germany"
names(de_l)[names(de_l) == "YEAR"] <- "Year"
names(de_l)[names(de_l) == "MONTH"] <- "Month"
names(de_l)[names(de_l) == "HOL_CODE"] <- "Replicate # (if raw data) from 1 to N"

## For now fixing sieve, year, month (still need method, samp area and zostera)
to_keep <- c("Month", "Year", "Site", "Country", "Mesh/sieve size (mm)",
             "Replicate # (if raw data) from 1 to N", "Sampling method",
             "Epifauna or Infauna", "Abundance", "Taxa", "Zostera dw (m-2)",
             "Zostera shoot density (m-2)", "`Zostera dw`", "Raw data or means?",
             "Sample size (m-2)", "Reference ID (from \"DataSources\" sheet)")

names(fin_both)[names(fin_both) %in% to_keep]
names(swe_both)[names(swe_both) %in% to_keep]
names(est_l)[names(est_l) %in% to_keep]
names(de_l)[names(de_l) %in% to_keep]
names(nor_both)[names(nor_both) %in% to_keep]


all_dat <- merge(fin_both[, names(fin_both) %in% to_keep],
                 swe_both[, names(swe_both) %in% to_keep],
                 all = TRUE)
all_dat <- merge(all_dat, nor_both[, names(nor_both) %in% to_keep], all = TRUE)
all_dat <- merge(all_dat, de_l[, names(de_l) %in% to_keep], all = TRUE)
all_dat <- merge(all_dat, est_l[, names(est_l) %in% to_keep], all = TRUE)


all_dat$zostera_dw_m2 <- all_dat$`Zostera dw (m-2)`
all_dat$zostera_shoots_m2 <- all_dat$`Zostera shoot density (m-2)`
all_dat$`Zostera dw (m-2)` <- NULL
all_dat$`Zostera shoot density (m-2)` <- NULL
all_dat$Sample_area_m2 <- as.numeric(all_dat$`Sample size (m-2)`)
all_dat$`Sample size (m-2)` <- NULL

all_dat[all_dat$Sampling_method %in% "mesh bag",
        "Sampling method"] <- "Mesh bag"
all_dat[all_dat$Sampling_method %in% "Other (specify in notes)",
        "Sampling method"] <- "Estonian method"

all_dat$Lit_ref <- all_dat$`Reference ID (from "DataSources" sheet)`
all_dat$`Reference ID (from "DataSources" sheet)` <- NULL


```


```{r}
## recoding specific variables

all_dat[all_dat$Month %in% c("April"), "Month"] <- 4
all_dat[all_dat$Month %in% c("August"), "Month"] <- 8
all_dat[all_dat$Month %in% c("August, 13"), "Month"] <- 8
all_dat[all_dat$Month %in% c("August, 20"), "Month"] <- 8
all_dat[all_dat$Month %in% c("August, 29"), "Month"] <- 8
all_dat[all_dat$Month %in% c("August,26"), "Month"] <- 8
all_dat[all_dat$Month %in% c("Augusti"), "Month"] <- 8
all_dat[all_dat$Month %in% c("December"), "Month"] <- 12
all_dat[all_dat$Month %in% c("November"), "Month"] <- 11
all_dat[all_dat$Month %in% c("October"), "Month"] <- 10
all_dat[all_dat$Month %in% c("Sept"), "Month"] <- 9
all_dat[all_dat$Month %in% c("September"), "Month"] <- 9
all_dat[all_dat$Month %in% c("Summer"), "Month"] <- 7
all_dat[all_dat$Month %in% c("summer"), "Month"] <- 7
all_dat[all_dat$Month %in% c("July"), "Month"] <- 7
all_dat[all_dat$Month %in% c("July, 29"), "Month"] <- 7
all_dat[all_dat$Month %in% c("July, mid"), "Month"] <- 7
all_dat[all_dat$Month %in% c("June"), "Month"] <- 6
all_dat[all_dat$Month %in% c("March"), "Month"] <- 3
all_dat[all_dat$Month %in% c("Mid ? September"), "Month"] <- 9
all_dat[all_dat$Month %in% c("August , 30"), "Month"] <- 8
all_dat[all_dat$Month %in% c("August , 2"), "Month"] <- 8
all_dat[all_dat$Month %in% c("August, 2"), "Month"] <- 8
all_dat[all_dat$Month %in% c("August , 29"), "Month"] <- 8
all_dat[all_dat$Month %in% c("August, 3"), "Month"] <- 8
all_dat[all_dat$Month %in% c("August, 30"), "Month"] <- 8
all_dat[all_dat$Month %in% c("August, 4"), "Month"] <- 8
all_dat[all_dat$Month %in% c("July, 23"), "Month"] <- 7
all_dat[all_dat$Month %in% c("September, 1"), "Month"] <- 9
all_dat[all_dat$Month %in% c("\"summer\""), "Month"] <- 7

all_dat$Month <- as.numeric(all_dat$Month)




## Remove wrong row
all_dat <- all_dat[!(all_dat$Site == "Hinderbengtsviken" &
                       all_dat$Year == 1993), ]

## Remove sites with less than 3 years
all_dat <- all_dat[all_dat$Month %in% c(6, 7, 8, 9), ]
all_dat_years <- all_dat[all_dat$Abundance > 0, ]
all_dat_years <- ddply(all_dat_years, c("Year", "Site"), summarise, N = length(unique(Month)))
three <- table(all_dat_years$Year, all_dat_years$Site)
to_keep <- c(colnames(three)[colSums(three > 0) >= 3])
all_dat <- all_dat[all_dat$Site %in% to_keep, ]

all_dat$epifauna_or_infauna <- all_dat$`Epifauna or Infauna`
all_dat$`Epifauna or Infauna` <- NULL

```


# Species look up and checking

After combining all the data to long format I produced a unique species list. To this species list I manually added genus.

- ignoring juveniles
- Add traits

```{r}
## Epifauna lookup
# species_names <- all_dat[, names(all_dat) %in% c("Taxa", "Country")]
# species_names <- dcast(species_names, Taxa ~ Country)
# species_names[, 2:6][species_names[, 2:6] > 0] <- 1
# 
# species_names$ScientificName <- species_names$Taxa
# species_names$Genus <- gsub(" .*", "", species_names$Taxa)
# species_names <- species_names[order(species_names$Taxa), ]

# write.csv2(species_names, row.names = FALSE, "data/epifauna_lookup.csv")


## Adding infauna
# infauna <- all_dat[all_dat$`Epifauna or Infauna` == "Infauna", ]
# species_names <- infauna[, names(infauna) %in% c("Taxa", "Country")]
# species_names <- dcast(species_names, Taxa ~ Country)
# species_names[, 2:4][species_names[, 2:4] > 0] <- 1
# 
# species_names$ScientificName <- species_names$Taxa
# species_names$Genus <- gsub(" .*", "", species_names$Taxa)
# species_names <- species_names[order(species_names$Taxa), ]
# lookup <- as.data.frame(read_xlsx("data/epifauna_lookup.xlsx"))
# species_names <- species_names[!species_names$Taxa %in% lookup$Taxa, ]
# write.csv(species_names, row.names = FALSE, "data/infauna_lookup.csv")
## Added to lookup
# library(worrms)
# 
# lookup <- as.data.frame(read_xlsx("data/species_lookup.xlsx"))
# 
# all_sp <- NULL
# for(i in 613:nrow(lookup)) {
#   print(i)
#   test <- wm_records_taxamatch(lookup$ScientificName[i], marine_only = FALSE)
#   all_sp <- rbind(all_sp, as.data.frame(test[[1]]))
# 
# }
# 
# write.csv2(all_sp, row.names = FALSE, "data/worms_checking.csv")
# 
# 
# all_sp <- read.csv2("data/worms_checking_ed.csv")
# all_sp <- all_sp[, names(all_sp) %in% c("AphiaID", "scientificname",
#                                         "valid_AphiaID", "valid_name",
#                                         "kingdom", "phylum", "class",
#                                         "order", "family", "genus")]
# 
# all_sp <- unique(all_sp)
# names(all_sp)[names(all_sp) == "scientificname"] <- "ScientificName"
# lookup_edit <- merge(lookup, all_sp, all.x = TRUE, by = "ScientificName")
# 
# write.csv2(lookup_edit, row.names = FALSE, "data/species_lookup.csv")
```


Adds row to lookup saying which species is where. This is necessary as some of the species might have dropped from the original when made requirments stricter

```{r}
# lookup <- as.data.frame(read_xlsx("data/species_lookup_new_1504.xlsx"))
# epifauna_species <- unique(all_dat[all_dat$`Epifauna or Infauna` %in% c("Epifauna"), "Taxa"])
# lookup[lookup$Taxa %in% epifauna_species, "in_epifauna"] <- "x"
# 
# epifauna_species <- unique(all_dat[all_dat$`Epifauna or Infauna` %in% c("Infauna"), "Taxa"])
# lookup[lookup$Taxa %in% epifauna_species, "in_infauna"] <- "x"
# 
# 
# epifauna_species <- unique(all_dat[all_dat$`Epifauna or Infauna` %in% c("Combined"), "Taxa"])
# lookup[lookup$Taxa %in% epifauna_species, "in_combined"] <- "x"
# 
# write.csv2(lookup, row.names = FALSE, "data/check_species.csv")


```



```{r}
#####
# lookup <- as.data.frame(read_xlsx("data/species_lookup_new.xlsx"))
lookup <- as.data.frame(read_xlsx("data/species_lookup_new_1504.xlsx"))

## Epifauna
fauna_lookup <- lookup[lookup$in_epifauna %in% "x" |
                         lookup$in_infauna %in% "x" |
                         lookup$in_combined %in% "x", ]

fauna_lookup <- fauna_lookup[, names(fauna_lookup) %in% c("Taxa", "ScientificName",
                                                          "genus", "Group", "valid_name",
                                                          "order", "class", "family", "phylum",
                                                          "Richness_epifauna", "Richness_infauna",
                                                          "in_infauna", "in_epifauna", "in_combined",
                                                          "DROP")]
table(fauna_lookup$Group)

dat_both <- merge(all_dat, fauna_lookup, by = "Taxa", all.x = TRUE)


dat_both$Replicate <- dat_both$`Replicate # (if raw data) from 1 to N`
dat_both$Sampling_method <- dat_both$`Sampling method`
dat_both$mesh_sieve_mm <- dat_both$`Mesh/sieve size (mm)`
dat_both$data_type <- dat_both$`Raw data or means?`
dat_both[dat_both$Country %in% "Germany", "data_type"] <- "raw"
dat_both$ID <- interaction(dat_both$Site, dat_both$Year, dat_both$Month)

## Merge per group
all_dat <- ddply(dat_both, c("Group", "Country", "Year", "ID",
                             "Month", "Site", "Replicate",
                             "epifauna_or_infauna", "genus", "ScientificName",
                             "zostera_dw_m2", "zostera_shoots_m2",
                             "mesh_sieve_mm", "Sampling_method",
                             "data_type", "Sample_area_m2", "valid_name",
                             "Lit_ref", "order", "class", "family", "phylum",
                             "Richness_epifauna", "Richness_infauna"),
                 summarise,
                 Abundance = sum(Abundance, na.rm = TRUE))

save(all_dat, file = "output/infauna_epifauna_042025.RData")


## Merge per group
all_dat_g <- ddply(dat_both, c("Group", "Country", "Year", "ID",
                               "Month", "Site", "Replicate", "Sample_area_m2",
                               "epifauna_or_infauna", "zostera_dw_m2", "zostera_shoots_m2",
                               "mesh_sieve_mm", "Sampling_method", "data_type",
                               "Richness_epifauna", "Richness_infauna"),
                   summarise,
                   Abundance = sum(Abundance, na.rm = TRUE))


save(all_dat_g, file = "output/infauna_epifauna_merge_group_042025.RData")


```

# Epifauna graph

```{r}
options(scipen = 999)

dat_plot <- all_dat_g[all_dat_g$Group %in% c("amphipoda", "bivalvia", "chironomidae", "decapoda", "gastropoda",
                                             "isopoda", "mysida", "polychaeta", "tanaidacea"), ]

dat_plot <- ddply(dat_plot, c("Group", "Country", "Year",
                              "Month", "Site", "epifauna_or_infauna"),
                  summarise,
                  Abundance = mean(Abundance, na.rm = TRUE))

dat_plot <- ddply(dat_plot, c("Group", "Country", "Year", "epifauna_or_infauna"), summarise,
                  Abundance = mean(Abundance, na.rm = TRUE))


plot_epi <- dat_plot[dat_plot$epifauna_or_infauna %in% c("Epifauna", "Combined"), ]

g <- ggplot(plot_epi, aes(Abundance, as.factor(Year), fill = Group))
g <- g + facet_wrap(~Country, drop = TRUE, scales = "free_y")
g <- g + xlab("% of total abundance") + ylab("Year")
g <- g + geom_bar(stat = "identity", position = "fill")
g

ggsave(g, file = "graphs/epifauna_relative_abundance_04.png", width = 10, height = 7.3)

g <- ggplot(plot_epi, aes(Abundance, as.factor(Year), fill = Group))
g <- g + facet_wrap(~Country, drop = TRUE, scales = "free")
g <- g + xlab("Abundance m2") + ylab("Year")
g <- g + geom_bar(stat = "identity", position = "dodge")
g

ggsave(g, file = "graphs/epifauna_abundance_grouped_04.png", width = 10, height = 7.3)


g <- ggplot(plot_epi, aes(Abundance, as.factor(Year), fill = Group))
g <- g + facet_grid(Country~Group, drop = TRUE, scales = "free")
g <- g + xlab("Abundance m2") + ylab("Year")
g <- g + geom_bar(stat = "identity")
g

ggsave(g, file = "graphs/epifauna_abundance_04.png", width = 25, height = 10)



```

# Infauna graph

```{r}

dat_plot <- all_dat_g[all_dat_g$Group %in% c("amphipoda", "bivalvia", "oligochaeta", "polychaeta"), ]

dat_plot <- ddply(dat_plot, c("Group", "Country", "Year",
                              "Month", "Site", "epifauna_or_infauna"),
                  summarise,
                  Abundance = mean(Abundance, na.rm = TRUE))

dat_plot <- ddply(dat_plot, c("Group", "Country", "Year", "epifauna_or_infauna"), summarise,
                  Abundance = mean(Abundance, na.rm = TRUE))

plot_in <- dat_plot[dat_plot$epifauna_or_infauna %in% c("Infauna", "Combined"), ]

g <- ggplot(plot_in, aes(Abundance, as.factor(Year), fill = Group))
g <- g + facet_wrap(~Country, drop = TRUE, scales = "free_y")
g <- g + xlab("% of total abundance") + ylab("Year")
g <- g + geom_bar(stat = "identity", position = "fill")
g

ggsave(g, file = "graphs/infauna_relative_abundance_04.png", width = 10, height = 7.3)

g <- ggplot(plot_in, aes(Abundance, as.factor(Year), fill = Group))
g <- g + facet_wrap(~Country, drop = TRUE, scales = "free")
g <- g + xlab("Abundance m2") + ylab("Year")
g <- g + geom_bar(stat = "identity", position = "dodge")
g

ggsave(g, file = "graphs/infauna_abundance_grouped_04.png", width = 10, height = 7.3)


g <- ggplot(plot_in, aes(Abundance, as.factor(Year), fill = Group))
g <- g + facet_grid(Country~Group, drop = TRUE, scales = "free")
g <- g + xlab("Abundance m2") + ylab("Year")
g <- g + geom_bar(stat = "identity")
g

ggsave(g, file = "graphs/infauna_abundance_04.png", width = 25, height = 10)
```


## Add coordinates

```{r}

coordinates <- read_xlsx("data/summeries/WP2 NORSE Data Status Oct.23_2024 .xlsx")
coordinates$Lat <- as.numeric(coordinates$Lat)
coordinates$Long <- as.numeric(coordinates$Long)
coordinates <- coordinates[!is.na(coordinates$Lat), ]
coordinates <- coordinates[!is.na(coordinates$Long), ]

# unique(all_dat_g$Site)[unique(all_dat_g$Site) %in% unique(coordinates$Site)]

sites_coords <- unique(coordinates[, names(coordinates) %in% c("Site",
                                                               "Lat",
                                                               "Long")])


sites_coords <- rbind(sites_coords,
                      data.frame(Site = "Gåsö", Lat = 58.233, Long = 11.400))

sites_coords_m <- ddply(sites_coords, c("Site"), summarise,
                        Lat = mean(Lat),
                        Long = mean(Long))




```


## Add SWM

```{r}

dat_sf <- st_as_sf(sites_coords_m, coords = c("Long", "Lat"), crs = 4326)
write_sf(dat_sf, "data/summeries/points.gpkg")

##
swm <- rast("V:/Arkisto VELMU1/Vanhat tai alkuperäiset aineistot/Taustamuuttujarasterit/Pinnan ekspositio/Expositio_baltic_wgs84/SWM_hela_Östersjön_WGS84/swm_with_wh1/sta.adf")

sites_coords_m$swm <- extract(swm, vect(dat_sf))$swm_with_wh1

## Salinity
sal <- rast("t:/Paperit/Remane/input/salinity_annual_0_10_13_23.tiff")
sal <- project(sal, crs(dat_sf))

sites_coords_m$salinity <- extract(sal, vect(dat_sf))$salinity_annual_0_10_13_23

sites_coords_m[sites_coords_m$Site == "Eru", "salinity"] <- 4.8
sites_coords_m[sites_coords_m$Site == "Tromper Wiek", "salinity"] <- 12.5
sites_coords_m[sites_coords_m$Site == "Sødra Stridsfjorden", "salinity"] <- 23.4
sites_coords_m[sites_coords_m$Site == "Kämpersvik", "salinity"] <- 23
sites_coords_m[sites_coords_m$Site == "Tromper Wiek", "salinity"] <- 12.5
sites_coords_m[sites_coords_m$Site == "Finsbo", "salinity"] <- 19


## SWM
sites_coords_m[sites_coords_m$Site == "Langerumpa", "swm"] <- 0.27
sites_coords_m[sites_coords_m$Site == "Klauva", "swm"] <- 0.07
sites_coords_m[sites_coords_m$Site == "Langerumpa", "salinity"] <- 26.6
sites_coords_m[sites_coords_m$Site == "Klauva", "salinity"] <- 12.6

## Unit nm2s1

all_dat_env <- merge(all_dat, sites_coords_m, all.x = TRUE)
all_dat_g_env <- merge(all_dat_g, sites_coords_m, all.x = TRUE)


save(all_dat_env, file = "output/infauna_epifauna_analysis_042025.RData")
save(all_dat_g_env, file = "output/infauna_epifauna_merge_group_analysis_042025.RData")

```


## Temperature

Temperature data was extracted from Copernicus for the 1982-2023. To calculate the temperature for each site we used the mean from a 15km area. To estimate the missing values we fit a gam to the timeseries and predicted the values for 2024 and years prior to 1982

```{r}
# sst <- rast("../Data/Copernicus/sst_subset.nc")
# sst_time <- time(sst)
# sst_time_month <- as.numeric(substr(sst_time, 6, 7))
# 
# temp_summer <- sst[[sst_time_month %in% c(6, 7, 8, 9)]]
# 
# years <- as.numeric(substr(time(temp_summer), 1, 4))
# 
# ## Make annual rasters
# for (i in 1:length(unique(years))) {
#   year <- unique(years)[[i]]
#   r_index <- years %in% year
#   
#   annual_r <- temp_summer[[r_index]]
#   annual_r <- mean(annual_r, na.rm = TRUE)
#   annual_r <- annual_r - 273.15
#   names(annual_r) <- year
#   
#   writeRaster(annual_r, filename = paste0("data/temperature/SST_", year,  ".tif"))
#   print(year)
#   
# }

dat_coords_sf <- st_as_sf(sites_coords_m, coords = c("Long", "Lat"), crs = 4326)

annual_temp <- rast(list.files("data/temperature", full.names = TRUE))
dat_coords_sf <- st_transform(dat_coords_sf, crs = st_crs(annual_temp))
dat_coords_sf_buff <- st_buffer(dat_coords_sf, dist = 15000)

temp_sites <- extract(annual_temp, vect(dat_coords_sf_buff),
                      mean, na.rm = TRUE)

temp_sites$Site <- dat_coords_sf_buff$Site

temp_sites_l <- melt(temp_sites, id.vars = c("Site", "ID"), variable.name = "Year",
                     value.name = "Temperature")
temp_sites_l$ID <- NULL
temp_sites_l$Year <- as.numeric(as.character(temp_sites_l$Year))

country <- all_dat_g_env[, names(all_dat_g_env) %in% c("Country", "Site")]
temp_sites_l <- merge(temp_sites_l, country, by = "Site")

g <- ggplot(temp_sites_l, aes(Year, Temperature, col = Site)) + geom_smooth()
g <- g + facet_wrap(~Country)
g


```

## Temperature model

```{r}

library(mgcv)

temp_sites_l$Site <- as.factor(temp_sites_l$Site)
temp_sites_l$Country <- as.factor(temp_sites_l$Country)

# mdl <- gam(Temperature ~ s(Year, k = 5, m = 2) +
#              s(Year, Country, k = 5, bs = "fs", m = 2),
#            data = temp_sites_l)
# 
# plot(mdl)
# gam.check(mdl)

mdl <- glm(Temperature ~ Year + Country + Site,
           data = temp_sites_l)

pred_frame <- expand.grid(Year = c(1969:1981, 2024),
                          Site = unique(temp_sites_l$Site))

pred_frame <- merge(pred_frame, country, by = "Site")
pred_frame <- pred_frame[!duplicated(pred_frame), ]

pred_frame$Temperature <- predict(mdl, pred_frame)

temp_full <- rbind(pred_frame, temp_sites_l)

```


```{r}

g <- ggplot(temp_full, aes(Year, Temperature, col = Site)) + geom_line()
g <- g + facet_wrap(~Country)
g

```

```{r}
## Add temperature
all_years_sites <- unique(all_dat_env[, names(all_dat_env) %in% c("Year", "Site")])
temp_sites <- unique(temp_full[paste0(temp_full$Site, temp_full$Year) %in%
                                 paste0(all_years_sites$Site, all_years_sites$Year), ])
temp_sites$Country <- NULL

all_dat_env_t <- merge(all_dat_env, temp_sites, all.x = TRUE, by = c("Year", "Site"))
all_dat_g_env_t <- merge(all_dat_g_env, temp_sites, all.x = TRUE, by = c("Year", "Site"))


save(all_dat_env_t, file = "output/infauna_epifauna_analysis_042025.RData")
save(all_dat_g_env_t, file = "output/infauna_epifauna_merge_group_analysis_042025.RData")


```





