---
title: "Epifauna analysis"
output: html_notebook
editor_options: 
  chunk_output_type: console
---



# Select data

```{r}
library(ggplot2)
library(reshape2)
library(plyr)
library(corrplot)
library(tidyverse)

#setwd("D:/Users/e1002963/Local/NORSE")

load("Epifauna_analysis_data.RData")

```
```{r fig.cap="Distribution of observations from all sites over time. Size indicates maximum number of groups observed that year."}

BD <- all_dat_ed[all_dat_ed$Abundance > 0, ]
BD <- ddply(BD, c("Year", "Site"), summarise, N = length(unique(Group)))

g <- ggplot(BD, aes(Year, Site, size = N)) + geom_point()
g <- g + scale_size(breaks = c(1, 5, 9))
g <- g + ylab("")
g <- g + theme_classic()
g


three <- table(all_dat_ed$Year, all_dat_ed$Site)
to_keep <- colnames(three)[colSums(three >= 3) >= 3]

dat <- all_dat_ed[all_dat_ed$Site %in% to_keep, ]

```

For the initial analysis we chose sites with more than 3 years with data, resulting in `r length(to_keep)` sites.


```{r fig.cap="Distribution of locations with min. 3 years of data."}
library(ggrepel)
sites <- dat %>%
  dplyr::select(Long, Lat, Site) %>% 
  filter(!is.na(Long)) %>%
  unique()

# Gåsø:
# 58.233/11.400


g <- ggplot(sites, aes(Long, Lat)) + geom_point() 
g <- g + geom_label_repel(aes(label = Site),
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50')
g <- g + borders(xlim = range(dat$Long, na.rm=T),ylim = range(dat$Lat, na.rm=T))
g <- g + theme_classic()
g


```


# Analysing differences in species occurrence

## Model & validation

```{r}
#### HMSC
library(Hmsc)

variables <- c("Site", "Country", "Year", "Month", "Lat", "Long", "swm", "salinity")
dat_w <- dcast(dat, Site + Country + Year + Month + Lat + Long + swm + salinity ~ Group,
               value.var = "Abundance", fill = 0)

dat_w <- dat_w[!dat_w$Site == "Gåsö", ]
dat_w$Month <- as.numeric(gsub("summer", 8, dat_w$Month))

species <- colnames(dat_w)[!names(dat_w) %in% variables]
Y <- dat_w[, names(dat_w) %in% species]

##
dat_w$Site <- as.factor(dat_w$Site)
dat_w$Country <- as.factor(dat_w$Country)
# dat_w$Replicate <- as.factor(dat_w$Replicate)

XData <- dat_w[, c("Year", "Site", "Month", "Country", "salinity")]

## Study design
studyDesign <- matrix(NA, nrow(Y), ncol = 2)
studyDesign[, 1] <- paste0("sample_", as.factor(1:nrow(dat_w))) # Sample
studyDesign[, 2] <- dat_w$Site
# studyDesign[, 3] <- dat_w$Country

studyDesign = as.data.frame(studyDesign)
colnames(studyDesign) = c("Sample","Site")
studyDesign[,1] = as.factor(studyDesign[,1])
studyDesign[,2] = as.factor(studyDesign[,2])
# studyDesign[,3] = as.factor(studyDesign[,3])

#Set random effects
rL1 = HmscRandomLevel(units = unique(studyDesign$Sample))
rL2 = HmscRandomLevel(units = unique(studyDesign$Site))
# rL3 = HmscRandomLevel(units = unique(studyDesign$Country))

xform <- ~Year + Month + Country + salinity# + swm

m = Hmsc(Y = 1 * (Y > 0), XData = XData, XFormula = xform,
         studyDesign = studyDesign,
         ranLevels = list(Sample = rL1, Site = rL2) ,
         distr = "probit")

thin <- 10
samples <- 500
transient <- 0.5 * (samples * thin)

# m = sampleMcmc(m, thin = thin, samples = samples, transient = transient,
#                nChains = 2, verbose = 0)

##

#load("epifauna_model_validation.RData")

##
```


```{r}

mpost = convertToCodaObject(m)
# effectiveSize(mpost$Beta)
# gelman.diag(mpost$Beta, multivariate = FALSE)$psrf
par(mfrow = c(2,2))
hist(effectiveSize(mpost$Beta), main="ess(beta)")
hist(gelman.diag(mpost$Beta, multivariate=FALSE)$psrf, main="psrf(beta)")
hist(effectiveSize(mpost$Omega[[1]]), main="ess(omega)")
hist(gelman.diag(mpost$Omega[[1]], multivariate=FALSE)$psrf, main="psrf(omega)")

```


```{r}
partition = createPartition(m, nfolds = 10)
preds = computePredictedValues(m, partition = partition)

evaluateModelFit(hM = m, predY = preds)

save(preds, m, file = "output/epifauna_pa_model_validation.RData")

```

## Graphs

```{r}
load("epifauna_pa_model_validation.RData")
## Plot Beta
postBeta = getPostEstimate(m, parName = "Beta")
plotBeta(m, post = postBeta, param = "Support",
         supportLevel = 0.95)

```


```{r}

OmegaCor = computeAssociations(m)
supportLevel = 0.95
toPlot = ((OmegaCor[[1]]$support > supportLevel)
          + (OmegaCor[[1]]$support < (1 - supportLevel)) > 0) * OmegaCor[[1]]$mean
corrplot(toPlot, method = "color",
         col = colorRampPalette(c("blue", "white", "red"))(200),
         title = paste("random effect level:", m$rLNames[1]),
         mar = c(0, 0, 1, 0))


```


```{r}

toPlot = ((OmegaCor[[2]]$support>supportLevel)
          + (OmegaCor[[2]]$support<(1-supportLevel))>0)*OmegaCor[[2]]$mean
corrplot(toPlot, method = "color",
         col = colorRampPalette(c("blue","white","red"))(200),
         title = paste("random effect level:", m$rLNames[2]), mar=c(0,0,1,0))

```


```{r}
group <- c(1, 1, 2, 3, 3, 3, 3, 3, 4)
groupnames = c("Year","Month","Country", "Salinity")
#Compute and plot VP
varp = computeVariancePartitioning(hM = m, group = group,
                                   groupnames = groupnames)
plotVariancePartitioning(m, varp)

```


```{r}
####

Gradient = constructGradient(m, focalVariable = "Year",
                             non.focalVariables = list(Site = list(3, "Tvärminne"),
                                                       Month = list(3, "August"),
                                                       Country = list(3, "Finland"),
                                                       Salinity = list(3, "Salinity")))
m$XData$Replicate <- NULL

predY = predict(m, XData = Gradient$XDataNew,
                studyDesign = Gradient$studyDesignNew,
                ranLevels = Gradient$rLNew, expected = TRUE)

plotGradient(m, Gradient, pred = predY, measure = "S", las = 1,
             showData = TRUE, main = 'Species richness (measure="S")')

```


## Analysing changes in species abundance

```{r}
#### HMSC

variables <- c("Site", "Country", "Year", "Month", "Lat", "Long", "swm", "salinity")
dat_w <- dcast(dat, Site + Country + Year + Month + Lat + Long + swm + salinity ~ Group,
               value.var = "Abundance", fill = 0)

dat_w <- dat_w[!dat_w$Site == "Gåsö", ]
dat_w$Month <- as.numeric(gsub("summer", 8, dat_w$Month))

species <- colnames(dat_w)[!names(dat_w) %in% variables]
Y <- dat_w[, names(dat_w) %in% species]

##
dat_w$Site <- as.factor(dat_w$Site)
dat_w$Country <- as.factor(dat_w$Country)
# dat_w$Replicate <- as.factor(dat_w$Replicate)

XData <- dat_w[, c("Year", "Site", "Month", "Country", "salinity")]

## Study design
studyDesign <- matrix(NA, nrow(Y), ncol = 2)
studyDesign[, 1] <- paste0("sample_", as.factor(1:nrow(dat_w))) # Sample
studyDesign[, 2] <- dat_w$Site
# studyDesign[, 3] <- dat_w$Country

studyDesign = as.data.frame(studyDesign)
colnames(studyDesign) = c("Sample","Site")
studyDesign[,1] = as.factor(studyDesign[,1])
studyDesign[,2] = as.factor(studyDesign[,2])
# studyDesign[,3] = as.factor(studyDesign[,3])

#Set random effects
rL1 = HmscRandomLevel(units = unique(studyDesign$Sample))
rL2 = HmscRandomLevel(units = unique(studyDesign$Site))
# rL3 = HmscRandomLevel(units = unique(studyDesign$Country))

xform <- ~Year + Month + Country + salinity# + swm

m = Hmsc(Y = Y, XData = XData, XFormula = xform,
         studyDesign = studyDesign,
         ranLevels = list(Sample = rL1, Site = rL2) ,
         distr = "poisson")

thin <- 10
samples <- 500
transient <- 0.5 * (samples * thin)

m = sampleMcmc(m, thin = thin, samples = samples, transient = transient,
               nChains = 2, verbose = 0)

##

# load("output/epifauna_model_validation.RData")

##
```


```{r}

mpost = convertToCodaObject(m)
# effectiveSize(mpost$Beta)
# gelman.diag(mpost$Beta, multivariate = FALSE)$psrf
par(mfrow = c(2,2))
hist(effectiveSize(mpost$Beta), main="ess(beta)")
hist(gelman.diag(mpost$Beta, multivariate=FALSE)$psrf, main="psrf(beta)")
hist(effectiveSize(mpost$Omega[[1]]), main="ess(omega)")
hist(gelman.diag(mpost$Omega[[1]], multivariate=FALSE)$psrf, main="psrf(omega)")

```


```{r}
partition = createPartition(m, nfolds = 10)
preds = computePredictedValues(m, partition = partition)

evaluateModelFit(hM = m, predY = preds)

save(preds, m, file = "output/epifauna_abund_model_validation.RData")

```


```{r}

load("epifauna_abund_model_validation.RData")
## Plot Beta
postBeta = getPostEstimate(m, parName = "Beta")
plotBeta(m, post = postBeta, param = "Support", supportLevel = 0.95)

```


```{r}
library(corrplot)
OmegaCor = computeAssociations(m)
supportLevel = 0.95
toPlot = ((OmegaCor[[1]]$support>supportLevel)
          + (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot(toPlot, method = "color",
         col = colorRampPalette(c("blue","white","red"))(200),
         title = paste("random effect level:", m$rLNames[1]), mar=c(0,0,1,0))
```


```{r}
toPlot = ((OmegaCor[[2]]$support>supportLevel)
          + (OmegaCor[[2]]$support<(1-supportLevel))>0)*OmegaCor[[2]]$mean
corrplot(toPlot, method = "color",
         col = colorRampPalette(c("blue","white","red"))(200),
         title = paste("random effect level:", m$rLNames[2]), mar=c(0,0,1,0))
```


```{r}

group <- c(1, 1, 2, 3, 3, 3, 3, 3, 4)
groupnames = c("Year","Month","Country", "Salinity")
#Compute and plot VP
varp = computeVariancePartitioning(hM = m, group = group,
                                   groupnames = groupnames)
plotVariancePartitioning(m, varp)

```


```{r}
####

Gradient = constructGradient(m, focalVariable = "Year",
                             non.focalVariables = list(Site = list(3, "Ängsö"),
                                                       Month = list(3, "August")))
m$XData$Replicate <- NULL

predY = predict(m, XData = Gradient$XDataNew,
                studyDesign = Gradient$studyDesignNew,
                ranLevels = Gradient$rLNew, expected = TRUE)

plotGradient(m, Gradient, pred = predY, measure = "S", las = 1,
             showData = TRUE, main = 'Species richness (measure="S")')

```

# Kristina's tests
Simple overviews of total abundance, species richness, diversity...
Site nested within country in the ANOVA. Germany stands out with higher abundance and diversity. 

```{r}
library(gridExtra)
library(vegan)


dat_overview <- dat %>%
  filter(!(Site == "Finsbo" & Year==2024)) %>% # Possible error, only zeroes
  mutate(across(c(Country, Site, Month), as.factor)) %>%
  group_by(Country, Site, Year, Month)  %>%
  summarise(TotAbundance = sum(Abundance),
    NumGroups = n_distinct(Group[Abundance > 0]),
    Shannon = diversity(Abundance, index = "shannon")) %>%
  ungroup()

m_abun <- aov(log(TotAbundance+1) ~ Country + Year + Month + Country:Site, data = dat_overview)
summary(m_abun)
# Significant effect of country

m_groups <- aov(NumGroups ~ Country + Year + Month + Country:Site, data = dat_overview)
summary(m_groups)
# Significant effect of country

m_div <- aov(Shannon ~ Country + Year + Month + Country:Site, data = dat_overview)
summary(m_div)
# Significant effect of country and site 

p1 <- ggplot(dat_overview, aes(x = Country, y = TotAbundance)) +
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x= NULL, y = "Abundance")

p2 <- ggplot(dat_overview, aes(x = Country, y = NumGroups)) +
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x= NULL, y = "Number of groups")

p3 <- ggplot(dat_overview, aes(x = Country, y = Shannon)) +
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = NULL, y = "Shannon-Wiener")

grid.arrange(p1, p2, p3, ncol = 3, bottom = "Country")

```

Overall trends with year, must be separated by country since data from Germany are relatively recent and these stand out. Month is never significant, removed

```{r}
library(mgcv)

# Create dataframe for model predictions
range_years <-  dat_overview  %>%
  group_by(Country)  %>%
  summarise(start_year = min(Year), end_year = max(Year))

newdat <- data.frame(Country = rep(unique(dat_overview$Country), each = 56),
                     Year = seq(min(dat_overview$Year),
                                      max(dat_overview$Year))) %>%
  left_join(range_years)  %>%
  group_by(Country)  %>%
  filter(Year >= start_year & Year <= end_year) %>%
  ungroup() %>%
  dplyr::select(Country, Year)

# GAM models
gam_abun <- gam(log(TotAbundance+1) ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_abun)
# No significant effects when using raw data
# After log transformation, significant year effects for Norway

gam_groups <- gam(NumGroups  ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_groups)
# Borderline for Norway  

gam_div <- gam(Shannon  ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_div)
# No significant effects

# Model predictions

newdat <- newdat  %>%
  mutate(TotAbundance = predict.gam(gam_abun, newdata = newdat),
         NumGroups = predict.gam(gam_groups, newdata = newdat),
         Shannon = predict.gam(gam_div, newdata = newdat))

p1 <- ggplot(dat_overview, aes(x = Year, y = log(TotAbundance+1), color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = TotAbundance, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "top") +
  labs(x= NULL, y = "log(Abundance)")

p2 <- ggplot(dat_overview, aes(x = Year, y = NumGroups, color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = NumGroups, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "none") +
 labs(x= NULL, y = "Number of groups")

p3 <- ggplot(dat_overview, aes(x = Year, y = Shannon, color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = Shannon, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "none") +
  labs(x = NULL, y = "Shannon-Wiener")

grid.arrange(p1, p2, p3, ncol = 1, bottom = "Year")

```

```{r}

dat_w <- dat %>% 
  filter(!(Site == "Finsbo" & Year==2024)) %>% # Possible error, only zeroes
  pivot_wider(names_from = Group, values_from = Abundance, values_fill = 0)  %>%   # Assuming missing value = 0
  arrange(Country)


comm_mat <- as.matrix(dat_w  %>%
  dplyr::select("polychaeta":"tanaidacea"))

dimnames(comm_mat) <- list(paste0(dat_w$Site,dat_w$Year), names(dat_w)[c(9:17)])

ord.fauna <- metaMDS(comm_mat)
stressplot(ord.fauna)

# Plot the NMDS with site names as text labels
# Grouped by country
countries <- unique(dat_w$Country)
cols <- rainbow(n = length(countries), alpha = 0.5)

plot(ord.fauna, type="n")
ordiellipse(ord.fauna, dat_w$Country, draw="polygon",label=FALSE, col = cols, lty =0)
orditorp(ord.fauna, type="n" ,display="sites",cex=0.7)
legend("bottomright", legend = countries, col = cols, pch = 19, bty = "n")

# Grouped by salinity level
salinity_levels <- c("Low (<10)","Medium (10-20)","High (>20)")
cols <- rainbow(n = length(salinity_levels), alpha = 0.5)

dat_w <- dat_w  %>%
  mutate(salinity_factor = case_when(
      salinity < 10 ~ "Low (<10)",
      salinity >= 10 & salinity < 20 ~ "Medium (10-20)",
      salinity >= 20 ~ "High (>20)")) %>%
  mutate(salinity_factor = factor(salinity_factor, levels = salinity_levels, ordered = TRUE))

plot(ord.fauna, type="n")
ordiellipse(ord.fauna, dat_w$salinity_factor, draw="polygon",label=FALSE, col = cols, lty =0)
orditorp(ord.fauna, type="n" ,display="sites",cex=0.7)
legend("bottomright", legend = salinity_levels, title = "Salinity", col = cols, pch = 19, bty = "n")


# Grouped by month
dat_w <- dat_w  %>%
  mutate(Month = if_else(Month == "summer", 7, as.numeric(Month)))

countries <- unique(dat_w$Country)
cols <- rainbow(n = length(countries), alpha = 0.5)

plot(ord.fauna, type="n")
ordiellipse(ord.fauna, dat_w$Month, draw="polygon",label=TRUE, col = cols, lty =0)
orditorp(ord.fauna, type="n" ,display="sites",cex=0.7)

```
