---
title: "Analyses for: Overview of biodiversity in Northern European seagrass meadows "
date: '`r format(Sys.Date(), "%d %B %Y")`'

output: 
  html_document:
    toc: true # Gives Table of Contents
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---



# Overview of data

```{r settings}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE) 

library(ggplot2)
library(reshape2)
library(plyr)
library(corrplot)
library(tidyverse)
library(knitr)
library(ggrepel)
library(scatterpie)
library(gridExtra)
library(vegan)
# Function to print tables
print_kable = function(x) {
  print(kable_print_output <<- x)
  cat('\n')
}

load("../../infauna_epifauna_merge_group_analysis_05032025.RData")  # Data grouped at class/order/family level
all_dat_g_env_t <- all_dat_g_env_t %>%
  mutate(Site = case_when(grepl("Finsbo",   Site) ~ "Finsbo", .default =  Site)) %>%
  # Site with missing long/lat (Nøtterøy) assumed to be 10.445 E/59.167 N
  mutate(Long = case_when(is.na(Long) ~ 10.445, .default =  Long)) %>%
  mutate(Lat = case_when(is.na(Long) ~ 59.167, .default =  Lat)) %>%
  mutate(Month = case_when(is.na(Month) ~ 8, .default =  Month))
  
load("../../infauna_epifauna_analysis_05032025.RData") # Data at genus/species level 
all_dat_env_t <- all_dat_env_t %>%
  mutate(Site = case_when(grepl("Finsbo",   Site) ~ "Finsbo", .default =  Site)) %>%
  # Site with missing long/lat (Nøtterøy) assumed to be 10.445 E/59.167 N
  mutate(Long = case_when(is.na(Long) ~ 10.445, .default =  Long)) %>%
  mutate(Lat = case_when(is.na(Long) ~ 59.167, .default =  Lat)) %>%
  mutate(Month = case_when(is.na(Month) ~ 8, .default =  Month))

```

Tables summarizing the data.

```{r table_data}
# Number of sites per country and range of years 
country_table <- all_dat_g_env_t %>%
  filter(Abundance > 0) %>%
  dplyr::select(Country, Site, Year, Set) %>%
  distinct() %>%
  group_by(Country, Set) %>%
  summarise(`Range years` =  paste0(min(Year),"-",max(Year)), `Num. sites` = length(unique(Site)))

kable(country_table, caption = "Number of sites with available eelgrass fauna data and range of years of data.")

# Summarizing the number of observations (years and months) per site
site_table <- all_dat_g_env_t %>%
  filter(Abundance > 0) %>%
  dplyr::select(Country, Site, Year, Month, Set) %>%
  distinct() %>%
  group_by(Country, Site) %>%
  summarise(`Range years` =  paste0(min(Year),"-",max(Year)), `Num. campaigns` = n(),`Type of data` = paste0(sort(unique(Set)), collapse=", "))
  
kable(site_table, caption = "All sites with available eelgrass fauna data, range of years of data and number of sampling campaigns for epifanua or infauna.")
rm(country_table, site_table)
```

Maps of the distribution of sites, including all sites or only sites with min. two years of data.

```{r map_all, fig.cap="Sites with eelgrass fauna data across the Northern European countries. The color scale of the label indicates the number of years of data available for each site. ", fig.height=10, fig.width=10}

sites <- all_dat_env_t %>%
  dplyr::select(Long, Lat, Site, Year, Set) %>% 
  unique() %>%
  group_by(Long, Lat, Site) %>%
  summarise(Num.years = length(unique(Year)), Fauna = paste0(sort(unique(Set)), collapse=","))  %>%
  mutate_at('Fauna', str_replace, 'Epifauna,Infauna', 'Both')

g <- ggplot(sites, aes(Long, Lat, shape = Fauna)) + geom_point() + 
  borders(xlim = range(sites$Long, na.rm=T),ylim = range(sites$Lat, na.rm=T)) + 
  theme_classic() + 
  geom_label_repel(data = sites,aes(Long, Lat, label = Site, color=Num.years),
                  box.padding   = 0.15, 
                  size=3,
                  max.overlaps = length(sites$Site),
                  segment.color = 'black') +
  scale_colour_gradient2(low = "grey",high = "black")
g

ggsave("../Graphs/map_all_sites.jpg", width=190, height=150,units="mm")
```

```{r map_selected, eval=FALSE, fig.cap="Sites minimum two years of eelgrass fauna data across the Northern European countries. The color scale of the label indicates the number of years of data available for each site.", fig.height=10, fig.width=10}

sites <- all_dat_env_t %>%
  dplyr::select(Long, Lat, Site, Year) %>% 
  unique() %>%
  group_by(Long, Lat, Site) %>%
  summarise(Num.years = n()) %>%
  filter(Num.years > 1)

g <- ggplot(sites, aes(Long, Lat)) + geom_point() + 
  borders(xlim = range(sites$Long, na.rm=T),ylim = range(sites$Lat, na.rm=T)) + 
  theme_classic() + 
  geom_label_repel(data = sites,aes(Long, Lat, label = Site, color=Num.years),
                  box.padding   = 0.15, 
                  size=3,
                  max.overlaps = length(sites$Site),
                  segment.color = 'black') +
  scale_colour_gradient2(low = "grey",high = "black")

g
ggsave("../Graphs/map_twoyears_sites.jpg", width=190, height=150,units="mm")
rm(sites)

```

Figures of the distributions of years and months of data

```{r fig_months_years, eval=TRUE, fig.cap="Distribution of observations from all sites over time. Size of the points indicate the number of months sampled that year, with color indicating which month.", fig.height=6, fig.width=10}

pie_data <- all_dat_env_t %>%
  mutate(Country_Site = paste(Country, Site, sep = ", "),
         Month = month.abb[as.numeric(Month)]) %>%
  dplyr::select(Country_Site, Year, Month) %>%
  distinct() %>%
  mutate(Value = 1)   %>%
  complete(Country_Site, Year, Month = month.abb, fill = list(Value = 0))  %>% 
  pivot_wider(names_from = Month, values_from = Value, values_fill = list(Value = 0)) %>% 
  dplyr::select(Country_Site, Year, month.abb) %>% 
  mutate(site_num = as.numeric(as.factor(Country_Site))) %>%
  mutate(count = rowSums(select(., contains(month.abb)))) 

countrysite2num <- pie_data %>%
  dplyr::select(Country_Site, site_num) %>%
  distinct() 

ggplot() + geom_scatterpie(data=pie_data, aes(x=Year, y=site_num, r = 0.5*log(count+1)),
                           cols=month.abb[1:12], color=NA) + 
  scale_y_continuous(breaks=countrysite2num$site_num, labels=countrysite2num$Country_Site) + 
  labs(x="Year", y="") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text=element_text(size=8),
        legend.key.size = unit(0.3, 'cm'), panel.background = element_rect(fill='white', colour='white'),
                    panel.grid.major = element_line(colour = "darkgrey"))

ggsave("../Graphs/piechart_year_month.jpg", width=190, height=120,units="mm")
rm(pie_data, countrysite2num)
```


# Analysing differences in biodiversity between countries

Simple overviews of total abundance, richness, diversity on the genus level.
Site nested within country in the ANOVA. Germany stands out with higher abundance and diversity. 

Trying the same but with genera.
```{r anovas_genera, eval=FALSE, fig.cap="Distribution of diversity indices of genera by country.", fig.height=10, fig.width=10}

plotlist <- list()
jitter_width <- 0.15
point_size <- 0.3

# Running model first for epifauna data, then infauna
for(set in c("Epifauna","Infauna")){
  dat_overview <- all_dat_env_t %>%
    # If genus info is missing, use ScientificName instead ?
    mutate(Genus = case_when(is.na(Genus) ~ ScientificName, .default =  Genus)) %>%
    filter(Set == set ) %>%
    mutate(across(c(Country, Site, Month), as.factor)) %>%
    # Average species abundances if multiple  replicates within site
    mutate(Replicate = replace_na(Replicate, "1")) %>% # If no replicates, give same number
    group_by(Country,Site,Year,Month,Genus,ScientificName) %>% 
    summarise(Abundance = mean(Abundance)) %>%
    ungroup()  %>%
    filter(!is.na(Genus) & Abundance >0) %>%
    group_by(Country, Site, Year, Month, Genus)  %>%
    summarise(GenusAbundance = sum(Abundance)) %>% # Summing the total abundance per genus (if more than one)
    ungroup() %>%
    group_by(Country, Site, Year, Month)  %>%
    summarise(TotAbundance = sum(GenusAbundance),
      NumGenus = n_distinct(Genus),
      Shannon = diversity(GenusAbundance, index = "shannon")) %>%
    ungroup()

  print(paste0("Formulating ANOVAs for ", set))

  m_abun <- aov(log(TotAbundance+1) ~ Country + Year + Month + Country:Site, data = dat_overview)
  print(m_abun$call)
  print(summary(m_abun))

  m_genus <- aov(NumGenus ~ Country + Year + Month + Country:Site, data = dat_overview)
  print(m_genus$call)
  print(summary(m_genus))

  m_div <- aov(Shannon ~ Country + Year + Month + Country:Site, data = dat_overview)
  print(m_div$call)
  print(summary(m_div))

  p1 <- ggplot(dat_overview, aes(x = Country, y = log(TotAbundance+1))) +
    geom_boxplot(outliers = FALSE) +
    geom_jitter(width = jitter_width, size = point_size) + 
    theme_classic() + 
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x= NULL, y = "log(Abundance)") +
    ggtitle(set)
  
  p2 <- ggplot(dat_overview, aes(x = Country, y = NumGenus)) +
    geom_boxplot(outliers = FALSE) +
    geom_jitter(width = jitter_width, size = point_size) + 
    theme_classic() + 
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x= NULL, y = "Number of genera")
  
  p3 <- ggplot(dat_overview, aes(x = Country, y = Shannon)) +
   geom_boxplot(outliers = FALSE) +
    geom_jitter(width = jitter_width, size = point_size) + 
    theme_classic() + 
     theme(axis.text.x = element_text(angle = 90)) +
    labs(x = NULL, y = "Shannon-Wiener")
  
  plotlist <- append(plotlist, list(p1, p2, p3))
  }

g <- grid.arrange(grobs=plotlist, ncol = 3, bottom = "Country")
ggsave(g, file = "../Graphs/boxplots_diversity_genera.jpg", width=190, height=190,units="mm")
rm(list=ls(pattern="m_"))
rm(plotlist, p1, p2, p3)
```

NMDS on genus level for epifauna.
```{r nmds_epifauna_genus, eval=FALSE, fig.cap="NMDS plot epifauna genera.", fig.height=10, fig.width=10}
salinity_levels <- c("Low (<10)","Medium (10-15)","High (>15)")

# Reshape data
dat_w <- all_dat_env_t %>% 
# Creating factor of salinity level
  filter(!is.na(salinity))  %>%
  mutate(salinity_factor = case_when(
      salinity < 10 ~ salinity_levels[1],
      salinity >= 10 & salinity < 15 ~ salinity_levels[2],
      salinity >= 15 ~ salinity_levels[3])) %>%
  mutate(salinity_factor = factor(salinity_factor, levels = salinity_levels, ordered = TRUE)) %>%
  #Remove missing data, select relevant data
  filter(!(Site == "Finsbo" & Year==2024)) %>% 
  filter(!is.na(Genus) & Set == "Epifauna") %>%
  dplyr::select(!c(Group)) %>%
  # Average species abundances if multiple  replicates within site?
  mutate(Replicate = replace_na(Replicate, "1")) %>% # If no replicates, give same number
  group_by(Site,Country,Year,Month,Genus,ScientificName, swm, salinity_factor) %>% 
  summarise(Abundance = mean(Abundance)) %>%
  ungroup()  %>%
  # Sum if multiple species within same genus
  group_by(Site,Country,Year,Month,Genus,swm,salinity_factor) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Remove rows with only zeroes
  filter(Abundance > 0)  %>%
  # Convert to wide format
  pivot_wider(names_from = Genus, values_from = Abundance, values_fill = 0)  %>%   # Assuming missing value = 0
  arrange(Country) 

# Making community matrix
comm_mat <- as.matrix(dat_w  %>%
  dplyr::select(-(Site:salinity_factor)))

colnames(comm_mat) <- colnames(dat_w  %>% dplyr::select(-(Site:salinity_factor)))
rownames(comm_mat) <- paste(dat_w$Site,dat_w$Year,dat_w$Month,sep="_")

dists <-vegdist(comm_mat)
ord.fauna <- metaMDS(dists, trymax = 100)
stressplot(ord.fauna)

# Plot the NMDS with site names as text labels
# Grouped by country
countries <- unique(dat_w$Country)
cols_countries <- rainbow(n = length(countries), alpha = 0.5)

#jpeg("../Graphs/MNDS_epifauna_genera.jpg", width=190, height=190,units="mm",res=300)
plot(ord.fauna, type="n")
ordiellipse(ord.fauna, dat_w$Country, draw="polygon",label=FALSE, col = cols_countries, lty =0)
orditorp(ord.fauna, type="n" ,display="sites",cex=0.7)
legend("bottomright", legend = countries, col = cols_countries, pch = 19, bty = "n")
ordiellipse(ord.fauna, dat_w$salinity_factor, draw="polygon",label=TRUE, lty =1:3,col="lightgrey",alpha=0.5)
#dev.off()
```

NMDS for infauna. 
```{r nmds_infauna_genus, eval=FALSE, fig.cap="NMDS plot infauna genera.", fig.height=10, fig.width=10}

# Reshape data
dat_w <- all_dat_env_t %>% 
# Creating factor of salinity level
  filter(!is.na(salinity))  %>%
  mutate(salinity_factor = case_when(
      salinity < 10 ~ salinity_levels[1],
      salinity >= 10 & salinity < 15 ~ salinity_levels[2],
      salinity >= 15 ~ salinity_levels[3])) %>%
  mutate(salinity_factor = factor(salinity_factor, levels = salinity_levels, ordered = TRUE)) %>%
  #Remove missing data, select relevant data
  filter(!(Site == "Finsbo" & Year==2024)) %>% 
  filter(!is.na(Genus) & Set == "Infauna") %>%
  dplyr::select(!c(Group)) %>%
  # Average species abundances if multiple  replicates within site?
  mutate(Replicate = replace_na(Replicate, "1")) %>% # If no replicates, give same number
  group_by(Site,Country,Year,Month,Genus,ScientificName, swm, salinity_factor) %>% 
  summarise(Abundance = mean(Abundance)) %>%
  ungroup()  %>%
  # Sum if multiple species within same genus
  group_by(Site,Country,Year,Month,Genus,swm,salinity_factor) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Remove rows with only zeroes
  filter(Abundance > 0)  %>%
  # Convert to wide format
  pivot_wider(names_from = Genus, values_from = Abundance, values_fill = 0)  %>%   # Assuming missing value = 0
  arrange(Country) 

# Making community matrix
comm_mat <- as.matrix(dat_w  %>%
  dplyr::select(-(Site:salinity_factor)))

colnames(comm_mat) <- colnames(dat_w  %>% dplyr::select(-(Site:salinity_factor)))
rownames(comm_mat) <- paste(dat_w$Site,dat_w$Year,dat_w$Month,sep="_")

dists <-vegdist(comm_mat)
ord.fauna <- metaMDS(dists, trymax = 100)
stressplot(ord.fauna)

# Plot the NMDS with site names as text labels
# Grouped by country
countries <- unique(dat_w$Country)
cols_countries <- rainbow(n = length(countries), alpha = 0.5)
cols_salinity <- rainbow(n = length(salinity_levels), alpha = 0.5)

plot(ord.fauna, type="n")
ordiellipse(ord.fauna, dat_w$Country, draw="polygon",label=FALSE, col = cols_countries, lty =0)
orditorp(ord.fauna, type="n" ,display="sites",cex=0.7)
legend("bottomright", legend = countries, col = cols_countries, pch = 19, bty = "n")
ordiellipse(ord.fauna, dat_w$salinity_factor, draw="polygon",label=TRUE, lty =1:3,col="lightgrey",alpha=0.5)

```

Investigate dominant genera, focusing on last 20 yrs and summer months (June to September)

```{r heatmap_genus, eval=FALSE, fig.cap="Heat map of species abundance (log-transformed), only including sites with data collected after 2010.", fig.height=10, fig.width=10}

# Reshape data
dat_heat <- all_dat_env_t %>% 
  #Remove missing data, select relevant data
  filter(Year>2010 & Month %in% c(6:9)) %>% 
  filter(!(Site == "Finsbo" & Year==2024)) %>% 
  filter(!is.na(Genus) & Set == "Epifauna") %>%
  dplyr::select(!c(Group)) %>%
  # Average species abundances if multiple  replicates within site?
  mutate(Replicate = replace_na(Replicate, "1")) %>% # If no replicates, give same number
  group_by(Site,Country,Year,Month,Genus,ScientificName) %>% 
  summarise(Abundance = mean(Abundance)) %>%
  ungroup()  %>%
  # Sum if multiple species within same genus
  group_by(Site,Country,Year,Month,Genus) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Average values per country
  group_by(Country,Genus) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Remove rows with only zeroes
  filter(Abundance > 0)  %>%
  mutate(species = factor(Genus, levels = sort(unique(Genus))))

ggplot(dat_heat, aes(x = Country, y = Genus, fill = log(Abundance+1))) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(heat.colors(10))) +
  theme_minimal() +
  labs(title = "Genus abundance per country (log(N+1))", x = "Country", y = "Genus", fill = "lnAbundance")

```

```{r yeartrend_gam, eval=FALSE}
# Analyzing yearly trends using GAMs, not included
library(mgcv)

# Create dataframe for model predictions
range_years <-  dat_overview  %>%
  group_by(Country)  %>%
  summarise(start_year = min(Year), end_year = max(Year))

newdat <- data.frame(Country = rep(unique(dat_overview$Country), each = 56),
                     Year = seq(min(dat_overview$Year),
                                      max(dat_overview$Year))) %>%
  left_join(range_years)  %>%
  group_by(Country)  %>%
  filter(Year >= start_year & Year <= end_year) %>%
  ungroup() %>%
  dplyr::select(Country, Year)

# GAM models
gam_abun <- gam(log(TotAbundance+1) ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_abun)
# No significant effects when using raw data
# After log transformation, significant year effects for Norway

gam_groups <- gam(NumGroups  ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_groups)
# Borderline for Norway  

gam_div <- gam(Shannon  ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_div)
# No significant effects

# Model predictions

newdat <- newdat  %>%
  mutate(TotAbundance = predict.gam(gam_abun, newdata = newdat),
         NumGroups = predict.gam(gam_groups, newdata = newdat),
         Shannon = predict.gam(gam_div, newdata = newdat))

p1 <- ggplot(dat_overview, aes(x = Year, y = log(TotAbundance+1), color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = TotAbundance, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "top") +
  labs(x= NULL, y = "log(Abundance)")

p2 <- ggplot(dat_overview, aes(x = Year, y = NumGroups, color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = NumGroups, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "none") +
 labs(x= NULL, y = "Number of groups")

p3 <- ggplot(dat_overview, aes(x = Year, y = Shannon, color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = Shannon, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "none") +
  labs(x = NULL, y = "Shannon-Wiener")

grid.arrange(p1, p2, p3, ncol = 1, bottom = "Year")

```
