---
title: "Analyses for: Overview of biodiversity in Northern European seagrass meadows "
date: '`r format(Sys.Date(), "%d %B %Y")`'

output: 
  html_document:
    toc: true # Gives Table of Contents
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---



# Overview of data

```{r settings}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE) 


library(ggplot2)
library(sf)
library(reshape2)
library(plyr)
library(corrplot)
library(tidyverse)
library(knitr)
library(ggrepel)
library(scatterpie)
library(gridExtra)
library(vegan)
library(captioner)
library(ggspatial)
# Function to print tables
print_kable = function(x) {
  print(kable_print_output <<- x)
  cat('\n')
}

# Figure labeling
figs <- captioner(prefix="Figure")
map_all <- figs(name="map_all","Sites with eelgrass fauna data across the Northern European countries. The color scale of the label indicates the number of years of data available for each site.")
fig_months_years <- figs(name="fig_months_years","Distribution of observations from all sites over time. Size of the points indicate the number of months sampled that year, with color indicating which month.")
anovas_genera <- figs(name="anovas_genera","Distribution of diversity indices of genera by country.")
nmds_epifauna_genus<- figs(name="nmds_epifauna_genus","NMDS plot epifauna genera, showing grouping by country and salinity level.")
nmds_infauna_genus<- figs(name="nmds_infauna_genus","NMDS plot epifauna genera, showing grouping by country and salinity level.")
heatmap_genus_epifauna<- figs(name="heatmap_genus_epifauna","Heat map of epifauna genus abundance (log-transformed), only including sites with data collected after 2010.")
heatmap_genus_infauna<- figs(name="heatmap_genus_infauna","Heat map of infauna genus abundance (log-transformed), only including sites with data collected after 2010.")

salinity_levels <- c("Low (<10)","Medium (10-15)","High (>15)")

# load("../../infauna_epifauna_merge_group_analysis_05032025.RData")  # Data grouped at class/order/family level
# all_dat_g_env_t <- all_dat_g_env_t %>%
#   mutate(Site = case_when(grepl("Finsbo",   Site) ~ "Finsbo", .default =  Site)) %>%
#   # Site with missing long/lat (Nøtterøy) assumed to be 10.445 E/59.167 N
#   mutate(Long = case_when(is.na(Long) ~ 10.445, .default =  Long)) %>%
#   mutate(Lat = case_when(is.na(Long) ~ 59.167, .default =  Lat)) %>%
#   mutate(Month = case_when(is.na(Month) ~ 8, .default =  Month))  %>%
#    mutate(salinity_factor = case_when(
#       salinity < 10 ~ salinity_levels[1],
#       salinity >= 10 & salinity < 15 ~ salinity_levels[2],
#       salinity >= 15 ~ salinity_levels[3])) %>%
#   mutate(salinity_factor = factor(salinity_factor, levels = salinity_levels, ordered = TRUE)) 

load("../../infauna_epifauna_analysis_042025.RData") # Data at genus/species level 

all_dat_env_t <- all_dat_env_t %>%
   mutate(salinity_factor = case_when(
      salinity < 10 ~ salinity_levels[1],
      salinity >= 10 & salinity < 15 ~ salinity_levels[2],
      salinity >= 15 ~ salinity_levels[3])) %>%
  mutate(salinity_factor = factor(salinity_factor, levels = salinity_levels, ordered = TRUE)) 

dat_epifauna <- all_dat_env_t[all_dat_env_t$epifauna_or_infauna %in% c("Epifauna"), ]
dat_epifauna <- dat_epifauna[!is.na(dat_epifauna$Richness_epifauna), ]
dat_epifauna$Richness <- dat_epifauna$Richness_epifauna

dat_infauna <- all_dat_env_t[all_dat_env_t$epifauna_or_infauna %in% c("Infauna"), ]
dat_infauna <- dat_infauna[!is.na(dat_infauna$Richness_infauna), ]
dat_infauna$Richness <- dat_infauna$Richness_infauna

dat_combined <- all_dat_env_t[all_dat_env_t$epifauna_or_infauna %in% c("Combined"), ]
dat_combined <- dat_combined[!is.na(dat_combined$Richness_infauna), ]
dat_combined$Richness <- NA

dat_infauna[is.na(dat_infauna$Sampling_method), "Sampling_method"] <- "Core"

both <- rbind(dat_infauna, dat_epifauna)

```


```{r table_data}
# Number of sites per country and range of years 
country_table <- all_dat_g_env_t %>%
  filter(Abundance > 0) %>%
  dplyr::select(Country, Site, Year, Set) %>%
  distinct() %>%
  group_by(Country, Set) %>%
  summarise(`Range years` =  paste0(min(Year),"-",max(Year)), `Num. sites` = length(unique(Site)))

kable(country_table, caption = "Number of sites with available eelgrass fauna data and range of years of data.")

# Summarizing the number of observations (years and months) per site
site_table <- all_dat_g_env_t %>%
  filter(Abundance > 0) %>%
  dplyr::select(Country, Site, Year, Month, Set) %>%
  distinct() %>%
  group_by(Country, Site) %>%
  summarise(`Range years` =  paste0(min(Year),"-",max(Year)), `Num. campaigns` = n(),`Type of data` = paste0(sort(unique(Set)), collapse=", "))
  
kable(site_table, caption = "All sites with available eelgrass fauna data, range of years of data and number of sampling campaigns for epifanua or infauna.")
rm(country_table, site_table)
```


```{r map_all, fig.cap=map_all, fig.height=10, fig.width=10}

sites <- all_dat_env_t %>%
  dplyr::select(Long, Lat, Site, Year, epifauna_or_infauna) %>% 
  unique() %>%
  group_by(Long, Lat, Site) %>%
  summarise(Num.years = length(unique(Year)), Fauna = paste0(sort(unique(epifauna_or_infauna)), collapse=","))  %>%
  mutate(Fauna = if_else(Fauna == "Epifauna", Fauna, "Epifauna and infauna"))


eu <- read_sf("../../shapefiles/europe_100km.shp")
sites_sf <- st_as_sf(sites[, names(sites) %in% c("Long", "Lat")],
                       coords = c("Long", "Lat"), crs = 4326)
sites_sf <- st_transform(sites_sf, crs = st_crs(eu))
sites_buffer <- st_buffer(sites_sf, dist = 100000)
eu <- st_crop(eu, sites_buffer)


g <- ggplot(eu) + geom_sf(col = "black", fill = "grey90") +
  geom_sf(data = sites_sf, fill = "black", col = "black") + 
  coord_sf(expand = FALSE, crs = 3035) +
  annotation_scale(location = "br", width_hint = 0.5) +
  theme_classic() + theme(legend.position = c(0.5, 0.4),
                 legend.background = element_blank(),
                 panel.border = element_rect(color = "black", fill = NA),
                 legend.text = element_text(size = 16),
                 axis.text = element_text(size = 16),
                 legend.title = element_blank())
g




g <- ggplot(sites, aes(Long, Lat, shape = Fauna)) +
  borders(xlim = range(sites$Long, na.rm=T),ylim = range(sites$Lat, na.rm=T), fill="black") + 
  coord_sf(expand = FALSE, crs = 3035) +

  theme_classic() + 
  geom_point(color="cyan", size = 1.5) + 
  geom_label_repel(data = sites,aes(Long, Lat, label = Site, color=Num.years),
                  box.padding   = 0.4, 
                  label.padding   = 0.1, 
                  size=3,
                  max.overlaps = length(sites$Site),
                  segment.color = 'cyan') +
  scale_colour_gradient2(low = "grey",high = "black")
g

ggsave("../Graphs/map_all_sites.jpg", width=190, height=170,units="mm")
```

```{r map_selected, eval=FALSE, fig.cap="Sites minimum two years of eelgrass fauna data across the Northern European countries. The color scale of the label indicates the number of years of data available for each site.", fig.height=10, fig.width=10}

sites <- all_dat_env_t %>%
  dplyr::select(Long, Lat, Site, Year) %>% 
  unique() %>%
  group_by(Long, Lat, Site) %>%
  summarise(Num.years = n()) %>%
  filter(Num.years > 1)

g <- ggplot(sites, aes(Long, Lat)) + geom_point() + 
  borders(xlim = range(sites$Long, na.rm=T),ylim = range(sites$Lat, na.rm=T)) + 
  theme_classic() + 
  geom_label_repel(data = sites,aes(Long, Lat, label = Site, color=Num.years),
                  box.padding   = 0.15, 
                  size=3,
                  max.overlaps = length(sites$Site),
                  segment.color = 'black') +
  scale_colour_gradient2(low = "grey",high = "black")

g
ggsave("../Graphs/map_twoyears_sites.jpg", width=190, height=150,units="mm")
rm(sites)

```


```{r fig_months_years, eval=TRUE, fig.cap=fig_months_years, fig.height=6, fig.width=10}

pie_data <- all_dat_env_t %>%
  mutate(Country_Site = paste(Country, Site, sep = ", "),
         Month = month.abb[as.numeric(Month)]) %>%
  dplyr::select(Country_Site, Year, Month) %>%
  distinct() %>%
  mutate(Value = 1)   %>%
  complete(Country_Site, Year, Month = month.abb, fill = list(Value = 0))  %>% 
  pivot_wider(names_from = Month, values_from = Value, values_fill = list(Value = 0)) %>% 
  dplyr::select(Country_Site, Year, month.abb) %>% 
  mutate(site_num = as.numeric(as.factor(Country_Site))) %>%
  mutate(count = rowSums(select(., contains(month.abb)))) 

countrysite2num <- pie_data %>%
  dplyr::select(Country_Site, site_num) %>%
  distinct() 

ggplot() + geom_scatterpie(data=pie_data, aes(x=Year, y=site_num, r = 0.5*log(count+1)),
                           cols=month.abb[1:12], color=NA) + 
  scale_y_continuous(breaks=countrysite2num$site_num, labels=countrysite2num$Country_Site) + 
  labs(x="Year", y="") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text=element_text(size=8),
        legend.key.size = unit(0.3, 'cm'), panel.background = element_rect(fill='white', colour='white'),
                    panel.grid.major = element_line(colour = "darkgrey"))

ggsave("../Graphs/piechart_year_month.jpg", width=190, height=120,units="mm")
rm(pie_data, countrysite2num)
```


# Analysing differences in biodiversity between countries

Simple overviews of total abundance, richness, diversity on the genus level.
Site nested within country in the ANOVA. Germany stands out with higher abundance and diversity. 

```{r anovas_genera, eval=TRUE, fig.cap=anovas_genera, fig.height=10, fig.width=10}

plotlist <- list()
jitter_width <- 0.15
point_size <- 0.3

# Running model first for epifauna data, then infauna
for(set in c("Epifauna","Infauna")){
  dat_overview <- all_dat_env_t %>%
    # If genus info is missing, use ScientificName instead ?
    #mutate(Genus = case_when(is.na(Genus) ~ ScientificName, .default =  Genus)) %>%
    filter(Set == set ) %>%
    mutate(across(c(Country, Site, Month), as.factor)) %>%
    # Average species abundances if multiple  replicates within site
    mutate(Replicate = replace_na(Replicate, "1")) %>% # If no replicates, give same number
    group_by(Country,Site,Year,Month,Genus,ScientificName) %>% 
    summarise(Abundance = mean(Abundance)) %>%
    ungroup()  %>%
    filter(!is.na(Genus) & Abundance >0) %>%
    group_by(Country, Site, Year, Month, Genus)  %>%
    summarise(GenusAbundance = sum(Abundance)) %>% # Summing the total abundance per genus (if more than one)
    ungroup() %>%
    group_by(Country, Site, Year, Month)  %>%
    summarise(TotAbundance = sum(GenusAbundance),
      NumGenus = n_distinct(Genus),
      Shannon = diversity(GenusAbundance, index = "shannon")) %>%
    ungroup()

  print(paste0("Formulating ANOVAs for ", set))

  m_abun <- aov(log(TotAbundance+1) ~ Country + Year + Month + Country:Site, data = dat_overview)
  print(m_abun$call)
  print(summary(m_abun))

  m_genus <- aov(NumGenus ~ Country + Year + Month + Country:Site, data = dat_overview)
  print(m_genus$call)
  print(summary(m_genus))

  m_div <- aov(Shannon ~ Country + Year + Month + Country:Site, data = dat_overview)
  print(m_div$call)
  print(summary(m_div))

  p1 <- ggplot(dat_overview, aes(x = Country, y = log(TotAbundance+1))) +
    geom_boxplot(outliers = FALSE) +
    geom_jitter(width = jitter_width, size = point_size) + 
    theme_classic() + 
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x= NULL, y = "log(Abundance)") +
    ggtitle(set)
  
  p2 <- ggplot(dat_overview, aes(x = Country, y = NumGenus)) +
    geom_boxplot(outliers = FALSE) +
    geom_jitter(width = jitter_width, size = point_size) + 
    theme_classic() + 
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x= NULL, y = "Number of genera")
  
  p3 <- ggplot(dat_overview, aes(x = Country, y = Shannon)) +
   geom_boxplot(outliers = FALSE) +
    geom_jitter(width = jitter_width, size = point_size) + 
    theme_classic() + 
     theme(axis.text.x = element_text(angle = 90)) +
    labs(x = NULL, y = "Shannon-Wiener")
  
  plotlist <- append(plotlist, list(p1, p2, p3))
  }

g <- grid.arrange(grobs=plotlist, ncol = 3, bottom = "Country")
ggsave(g, file = "../Graphs/boxplots_diversity_genera.jpg", width=190, height=190,units="mm")
rm(list=ls(pattern="m_"))
rm(plotlist, p1, p2, p3)
```

NMDS on genus level for epifauna.
```{r nmds_epifauna_genus, eval=TRUE, fig.cap=nmds_epifauna_genus, fig.height=10, fig.width=10}

# Reshape data to wide format
dat_w_epi <- all_dat_env_t %>% 
  # If genus info is missing, use ScientificName instead ?
  #mutate(Genus = case_when(is.na(Genus) ~ ScientificName, .default =  Genus)) %>%
  #Remove missing data, select relevant data
  filter(!(Site == "Finsbo" & Year==2024)) %>% 
  filter(!is.na(Genus) & Set == "Epifauna" & !is.na(salinity_factor)) %>%
  dplyr::select(!c(Group)) %>%
  # Average species abundances if multiple  replicates within site
  mutate(Replicate = replace_na(Replicate, "1")) %>% # If no replicates, give same number
  group_by(Site,Country,Year,Month,Genus,ScientificName, swm, salinity_factor) %>% 
  summarise(Abundance = mean(Abundance)) %>%
  ungroup()  %>%
  # Sum if multiple species within same genus
  group_by(Site,Country,Year,Month,Genus,swm,salinity_factor) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Remove rows with only zeroes
  filter(Abundance > 0)  %>%
  # Convert to wide format
  pivot_wider(names_from = Genus, values_from = Abundance, values_fill = 0)  %>%   # Assuming missing value = 0
  arrange(Country) 

# Remove taxa only counted < X times?
threshold <- 0
#stress <- c()
#for(threshold in 1:20){
  
keep_taxa <- names(dat_w_epi  %>%
  dplyr::select(-(Site:salinity_factor)) %>%
  summarise(across(everything(), ~ sum(. > 0)))  %>%
  select(where(~ . > threshold)))

# Making community matrix
comm_mat_epi <- as.matrix(dat_w_epi  %>%
  dplyr::select(all_of(keep_taxa)))
  
colnames(comm_mat_epi) <- keep_taxa
rownames(comm_mat_epi) <- paste(dat_w_epi$Site,dat_w_epi$Year,dat_w_epi$Month,sep="_")

# Calculating distance matrix and NMDS
dists_epi <-vegdist(comm_mat_epi)
ord.epifauna <- metaMDS(dists_epi, trymax = 100)
#stressplot(ord.epifauna)
#stress <- c(stress,ord.epifauna$stress)
print(ord.epifauna)

#}
#plot(1:20, stress)

# https://rpubs.com/CPEL/NMDS
# As a rule of thumb literature has identified the following cut-off values for stress-level:
# Higher than 0.2 is poor (risks for false interpretation).
# 0.1 - 0.2 is fair (some distances can be misleading for interpretation).
# 0.05 - 0.1 is good (can be confident in inferences from plot).
# Less than 0.05 is excellent (this can be rare).

# Plot the NMDS with site names as text labels, grouped by country and salinity level
countries <- unique(dat_w_epi$Country)
cols_countries <- rainbow(n = length(countries), alpha = 0.5)

#jpeg("../Graphs/MNDS_epifauna_genera.jpg", width=190, height=170,units="mm",res=300)
plot(ord.epifauna, type="n")
ordiellipse(ord.epifauna, dat_w_epi$Country, draw="polygon",label=FALSE, col = cols_countries, lty =0)
orditorp(ord.epifauna, type="n" ,display="sites",cex=0.5)
legend("bottomleft", legend = countries, col = cols_countries, pch = 19, bty = "n")
ordiellipse(ord.epifauna, dat_w_epi$salinity_factor, draw="polygon",label=TRUE, lty =1:3,col="lightgrey",alpha=0.5, cex = 0.8) 
mtext(side = 3,paste('NMDS for epifauna, stress =',round(ord.epifauna$stress,3))) #add stress to plot
#dev.off()
```

```{r permanova_epifauna_genus, eval=TRUE}
# Perform PERMANOVA to investigate significance of factors
perm_epi <- adonis2(dists_epi ~ as.factor(Country) + as.factor(salinity_factor), data = dat_w_epi, permutations=9999, by ="terms")
print(perm_epi)

# SIMPER to identify taxa contributing to difference
sim_epi <- simper(comm_mat_epi, dat_w_epi$Country)

# check https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/simper/ 
# summary(sim_epi)$Germany_Norway %>%
#   round(3) %>%
#   head()

comparisons <- names(sim_epi)
comparisons_colnames <- gsub(" ",".",comparisons)

simper.results <- c()
for(i in 1:length(comparisons)) {
  temp <- summary(sim_epi)[as.character(comparisons[i])] %>%
    as.data.frame()
  colnames(temp) <- gsub(paste(comparisons_colnames[i],".", sep = ""), "", colnames(temp))
  temp <- temp %>%
    mutate(Comparison = comparisons[i],
    Position = row_number()) %>%
      rownames_to_column(var = "Genus")
  simper.results <- rbind(simper.results, temp)
}

simper_top <- simper.results %>%
  filter(Position == 1) %>%
  dplyr::select(Comparison, Genus, p)
  #filter(Position %in% c(1:3)) %>%
  #filter(p <= 0.05) %>%
  #dplyr::select(Comparison, Genus, Position, p)

print(simper_top)

```

NMDS for infauna. 
```{r nmds_infauna_genus, eval=TRUE, fig.cap=nmds_infauna_genus, fig.height=10, fig.width=10}

# Reshape data
dat_w_in <- all_dat_env_t %>% 
  #Remove missing data, select relevant data
  filter(!is.na(Genus) & !is.na(salinity_factor) & Set == "Infauna") %>%
  filter(!(Site == "Finsbo" & Year==2024)) %>% 
  dplyr::select(!c(Group)) %>%
  # Average species abundances if multiple  replicates within site?
  mutate(Replicate = replace_na(Replicate, "1")) %>% # If no replicates, give same number
  group_by(Site,Country,Year,Month,Genus,ScientificName, swm, salinity_factor) %>% 
  summarise(Abundance = mean(Abundance)) %>%
  ungroup()  %>%
  # Sum if multiple species within same genus
  group_by(Site,Country,Year,Month,Genus,swm,salinity_factor) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Remove rows with only zeroes
  filter(Abundance > 0)  %>%
  # Convert to wide format
  pivot_wider(names_from = Genus, values_from = Abundance, values_fill = 0)  %>%   # Assuming missing value = 0
  arrange(Country) 

# Making community matrix
comm_mat_in <- as.matrix(dat_w_in  %>%
  dplyr::select(-(Site:salinity_factor)))

colnames(comm_mat_in) <- colnames(dat_w_in  %>% dplyr::select(-(Site:salinity_factor)))
rownames(comm_mat_in) <- paste(dat_w_in$Site,dat_w_in$Year,dat_w_in$Month,sep="_")

dists_in <-vegdist(comm_mat_in)
ord.infauna <- metaMDS(dists_in, trymax = 100)
print(ord.infauna)
#stressplot(ord.infauna)

# Plot the NMDS with site names as text labels
# Grouped by country
countries <- unique(dat_w_in$Country)
cols_countries <- rainbow(n = length(countries), alpha = 0.5)
cols_salinity <- rainbow(n = length(salinity_levels), alpha = 0.5)

#jpeg("../Graphs/MNDS_infauna_genera.jpg", width=190, height=170,units="mm",res=300)
plot(ord.infauna, type="n")
ordiellipse(ord.infauna, dat_w_in$Country, draw="polygon",label=FALSE, col = cols_countries, lty =0)
orditorp(ord.infauna, display="sites",cex=0.7)
legend("bottomleft", legend = countries, col = cols_countries, pch = 19, bty = "n")
ordiellipse(ord.infauna, dat_w_in$salinity_factor, draw="polygon",label=TRUE, lty =1:3,col="lightgrey",alpha=0.5)
mtext(side = 3,paste('NMDS for infauna, stress =',round(ord.infauna$stress,3))) #add stress to plot
#dev.off()
```

```{r permanova_INfauna_genus, eval=TRUE}
# Perform PERMANOVA to investigate significance of factors
perm_in <- adonis2(dists_in ~ as.factor(Country) + as.factor(salinity_factor), data = dat_w_in, permutations=9999, by ="terms")
print(perm_in)

# SIMPER to identify taxa contributing to difference
sim_in <- simper(comm_mat_in, dat_w_in$Country)

comparisons <- names(sim_in)
comparisons_colnames <- gsub(" ",".",comparisons)

simper.results <- c()
for(i in 1:length(comparisons)) {
  temp <- summary(sim_in)[as.character(comparisons[i])] %>%
    as.data.frame()
  colnames(temp) <- gsub(paste(comparisons_colnames[i],".", sep = ""), "", colnames(temp))
  temp <- temp %>%
    mutate(Comparison = comparisons[i],
    Position = row_number()) %>%
      rownames_to_column(var = "Genus")
  simper.results <- rbind(simper.results, temp)
}

simper_top <- simper.results %>%
  filter(Position == 1) %>%
  dplyr::select(Comparison, Genus, p)
  #filter(Position %in% c(1:3)) %>%
  #filter(p <= 0.05) %>%
  #dplyr::select(Comparison, Genus, Position, p)

print(simper_top)

```

Investigate dominant genera, focusing on last 20 yrs and summer months (June to September)

```{r heatmap_genus_epifauna, eval=TRUE, fig.cap="Heat map of epifauna genus abundance (log-transformed), only including sites with data collected after 2010 and months 6-9.", fig.height=12, fig.width=10}

# Reshape data
dat_heat <- all_dat_env_t %>% 
  #Remove missing data, select relevant data
  filter(Year>2010 & Month %in% c(6:9)) %>% 
  filter(!(Site == "Finsbo" & Year==2024)) %>% 
  filter(!is.na(Genus) & Set == "Epifauna") %>%
  dplyr::select(!c(Group)) %>%
  # Average species abundances if multiple  replicates within site?
  mutate(Replicate = replace_na(Replicate, "1")) %>% # If no replicates, give same number
  group_by(Site,Country,Year,Month,Genus,ScientificName) %>% 
  summarise(Abundance = mean(Abundance)) %>%
  ungroup()  %>%
  # Sum if multiple species within same genus
  group_by(Site,Country,Year,Month,Genus) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Average values per country
  group_by(Country,Genus) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Remove rows with only zeroes
  filter(Abundance > 0)  %>%
  mutate(species = factor(Genus, levels = sort(unique(Genus))))

ggplot(dat_heat, aes(x = Country, y = Genus, fill = log(Abundance+1))) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(heat.colors(10))) +
  theme_minimal() +
  labs(title = "Genus abundance per country (log(N+1))", x = "Country", y = "Genus", fill = "lnAbundance")

```

```{r heatmap_genus_infauna, eval=TRUE, fig.cap="Heat map of infaunga genus abundance (log-transformed), only including sites with data collected after 2010 and months 6-9.", fig.height=12, fig.width=10}

# Reshape data
dat_heat <- all_dat_env_t %>% 
  #Remove missing data, select relevant data
  filter(Year>2010 & Month %in% c(6:9)) %>% 
  filter(!(Site == "Finsbo" & Year==2024)) %>% 
  filter(!is.na(Genus) & Set == "Infauna") %>%
  dplyr::select(!c(Group)) %>%
  # Average species abundances if multiple  replicates within site?
  mutate(Replicate = replace_na(Replicate, "1")) %>% # If no replicates, give same number
  group_by(Site,Country,Year,Month,Genus,ScientificName) %>% 
  summarise(Abundance = mean(Abundance)) %>%
  ungroup()  %>%
  # Sum if multiple species within same genus
  group_by(Site,Country,Year,Month,Genus) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Average values per country
  group_by(Country,Genus) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Remove rows with only zeroes
  filter(Abundance > 0)  %>%
  mutate(species = factor(Genus, levels = sort(unique(Genus))))

ggplot(dat_heat, aes(x = Country, y = Genus, fill = log(Abundance+1))) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(heat.colors(10))) +
  theme_minimal() +
  labs(title = "Genus abundance per country (log(N+1))", x = "Country", y = "Genus", fill = "lnAbundance")

```

```{r yeartrend_gam, eval=FALSE}
# Analyzing yearly trends using GAMs, not included
library(mgcv)

# Create dataframe for model predictions
range_years <-  dat_overview  %>%
  group_by(Country)  %>%
  summarise(start_year = min(Year), end_year = max(Year))

newdat <- data.frame(Country = rep(unique(dat_overview$Country), each = 56),
                     Year = seq(min(dat_overview$Year),
                                      max(dat_overview$Year))) %>%
  left_join(range_years)  %>%
  group_by(Country)  %>%
  filter(Year >= start_year & Year <= end_year) %>%
  ungroup() %>%
  dplyr::select(Country, Year)

# GAM models
gam_abun <- gam(log(TotAbundance+1) ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_abun)
# No significant effects when using raw data
# After log transformation, significant year effects for Norway

gam_groups <- gam(NumGroups  ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_groups)
# Borderline for Norway  

gam_div <- gam(Shannon  ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_div)
# No significant effects

# Model predictions

newdat <- newdat  %>%
  mutate(TotAbundance = predict.gam(gam_abun, newdata = newdat),
         NumGroups = predict.gam(gam_groups, newdata = newdat),
         Shannon = predict.gam(gam_div, newdata = newdat))

p1 <- ggplot(dat_overview, aes(x = Year, y = log(TotAbundance+1), color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = TotAbundance, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "top") +
  labs(x= NULL, y = "log(Abundance)")

p2 <- ggplot(dat_overview, aes(x = Year, y = NumGroups, color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = NumGroups, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "none") +
 labs(x= NULL, y = "Number of groups")

p3 <- ggplot(dat_overview, aes(x = Year, y = Shannon, color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = Shannon, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "none") +
  labs(x = NULL, y = "Shannon-Wiener")

grid.arrange(p1, p2, p3, ncol = 1, bottom = "Year")

```
