---
title: "Analyses for: Overview of biodiversity in Northern European seagrass meadows "
date: '`r format(Sys.Date(), "%d %B %Y")`'

output: 
  html_document:
    toc: true # Gives Table of Contents
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---



# Overview of data

```{r settings}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE) 

library(ggplot2)
library(reshape2)
library(plyr)
library(corrplot)
library(tidyverse)
library(knitr)
library(ggrepel)
library(scatterpie)
library(gridExtra)
library(vegan)
# Function to print tables
print_kable = function(x) {
  print(kable_print_output <<- x)
  cat('\n')
}

load("../../05 Data/infauna_epifauna_merge_group_analysis_24022025.RData") # Data grouped at class/order/family level
load("../../05 Data/infauna_epifauna_analysis_24022025.RData") # Data at genus/species level 

# Site with missing long/lat (Nøtterøy) assumed to be 10.445 E/59.167 N
all_dat_g_env$Long[is.na(all_dat_g_env$Long)] <- 10.445
all_dat_g_env$Lat[is.na(all_dat_g_env$Lat)] <- 59.167
all_dat_env$Long[is.na(all_dat_env$Long)] <- 10.445
all_dat_env$Lat[is.na(all_dat_env$Lat)] <- 59.167

```

Tables summarizing the data.

```{r table_data}
# Number of sites per country and range of years 
country_table <- all_dat_g_env %>%
  filter(Abundance > 0) %>%
  dplyr::select(Country, Site, Year) %>%
  distinct() %>%
  group_by(Country) %>%
  summarise(`Range years` =  paste0(min(Year),"-",max(Year)), `Num. sites` = length(unique(Site)))

kable(country_table, caption = "Number of sites with available eelgrass fauna data and range of years of data.")

# Summarizing the number of observations (years and months) per site
site_table <- all_dat_g_env %>%
  filter(Abundance > 0) %>%
  dplyr::select(Country, Site, Year, Month) %>%
  distinct() %>%
  group_by(Country, Site) %>%
  summarise(`Range years` =  paste0(min(Year),"-",max(Year)), `Num. campaigns` = n())
  
kable(site_table, caption = "All sites with available eelgrass fauna data, range of years of data and number of sampling campaigns for epifanua or infauna.")
rm(country_table, site_table)
```

Maps of the distribution of sites, including all sites or only sites with min. two years of data.

```{r map_all, fig.cap="Sites with eelgrass fauna data across the Northern European countries. The color scale of the label indicates the number of years of data available for each site. ", fig.height=10, fig.width=10}

sites <- all_dat_env %>%
  dplyr::select(Long, Lat, Site, Year) %>% 
  unique() %>%
  group_by(Long, Lat, Site) %>%
  summarise(Num.years = n()) 

g <- ggplot(sites, aes(Long, Lat)) + geom_point() + 
  borders(xlim = range(sites$Long, na.rm=T),ylim = range(sites$Lat, na.rm=T)) + 
  theme_classic() + 
  geom_label_repel(data = sites,aes(Long, Lat, label = Site, color=Num.years),
                  box.padding   = 0.15, 
                  size=3,
                  max.overlaps = length(sites$Site),
                  segment.color = 'black') +
  scale_colour_gradient2(low = "grey",high = "black")
g

ggsave("../Graphs/map_all_sites.jpg", width=190, height=150,units="mm")
```

```{r map_selected, fig.cap="Sites minimum two years of eelgrass fauna data across the Northern European countries. The color scale of the label indicates the number of years of data available for each site.", fig.height=10, fig.width=10}

sites <- all_dat_env %>%
  dplyr::select(Long, Lat, Site, Year) %>% 
  unique() %>%
  group_by(Long, Lat, Site) %>%
  summarise(Num.years = n()) %>%
  filter(Num.years > 1)

g <- ggplot(sites, aes(Long, Lat)) + geom_point() + 
  borders(xlim = range(sites$Long, na.rm=T),ylim = range(sites$Lat, na.rm=T)) + 
  theme_classic() + 
  geom_label_repel(data = sites,aes(Long, Lat, label = Site, color=Num.years),
                  box.padding   = 0.15, 
                  size=3,
                  max.overlaps = length(sites$Site),
                  segment.color = 'black') +
  scale_colour_gradient2(low = "grey",high = "black")

g
ggsave("../Graphs/map_twoyears_sites.jpg", width=190, height=150,units="mm")
rm(sites)

```

Figures of the distributions of years and months of data

```{r fig_months_uears, fig.cap="Distribution of observations from all sites over time. Size of the points indicate the number of months sampled that year, with color indicating which month.", fig.height=20, fig.width=10}

pie_data <- all_dat_env %>%
  filter(!is.na(Month)) %>%
  mutate(Country_Site = paste(Country, Site, sep = ", "),
         Month = month.abb[as.numeric(Month)]) %>%
  dplyr::select(Country_Site, Year, Month) %>%
  distinct() %>%
  mutate(Value = 1)   %>%
  complete(Country_Site, Year, Month = month.abb, fill = list(Value = 0))  %>% 
  pivot_wider(names_from = Month, values_from = Value, values_fill = list(Value = 0)) %>% 
  dplyr::select(Country_Site, Year, month.abb) %>% 
  mutate(site_num = as.numeric(as.factor(Country_Site))) %>%
  mutate(count = rowSums(select(., contains(month.abb)))) 

countrysite2num <- pie_data %>%
  dplyr::select(Country_Site, site_num) %>%
  distinct() 

ggplot() + geom_scatterpie(data=pie_data, aes(x=Year, y=site_num, r = 0.5*log(count+1)),
                           cols=month.abb[1:12], color=NA) + 
  scale_y_continuous(breaks=countrysite2num$site_num, labels=countrysite2num$Country_Site) + 
  labs(x="Year", y="") + 
  theme_classic() +
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text=element_text(size=8), legend.key.size = unit(0.3, 'cm'))

ggsave("../Graphs/piechart_year_month.jpg", width=190, height=100,units="mm")
rm(pie_data, countrysite2num)
```


# Analysing differences in biodiversity between countries

Simple overviews of total abundance, richness, diversity of oveall groups of epifauna.
Site nested within country in the ANOVA. Germany stands out with higher abundance and diversity. 

```{r anovas_epifauna_groups, eval=TRUE, fig.cap="Distribution of diversity indices of coarse groups by country.", fig.height=10, fig.width=10}

dat_overview <- all_dat_g_env %>%
  filter(Set == "Epifauna") %>%
  mutate(across(c(Country, Site, Month), as.factor)) %>%
  group_by(Country, Site, Year, Month, Group)  %>%
  summarise(GroupAbundance = sum(Abundance)) %>% # Summing the total abundance per group 
  ungroup() %>%
  filter(GroupAbundance > 0)  %>%
  group_by(Country, Site, Year, Month)  %>%
  summarise(TotAbundance = sum(GroupAbundance),
    NumGroups = n_distinct(Group),
    Shannon = diversity(GroupAbundance, index = "shannon")) %>%
  ungroup()

m_abun <- aov(log(TotAbundance+1) ~ Country + Year + Month + Country:Site, data = dat_overview)
print(m_abun$call)
print(summary(m_abun))
# Significant effect of country

m_groups <- aov(NumGroups ~ Country + Year + Month + Country:Site, data = dat_overview)
print(m_groups$call)
print(summary(m_groups))
# Significant effect of country

m_div <- aov(Shannon ~ Country + Year + Month + Country:Site, data = dat_overview)
print(m_div$call)
print(summary(m_div))
# Significant effect of country and site 

jitter_width <- 0.15
point_size <- 0.5

p1 <- ggplot(dat_overview, aes(x = Country, y = TotAbundance)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = jitter_width, size = point_size) + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x= NULL, y = "Abundance")

p2 <- ggplot(dat_overview, aes(x = Country, y = NumGroups)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = jitter_width, size = point_size) + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x= NULL, y = "Number of groups")

p3 <- ggplot(dat_overview, aes(x = Country, y = Shannon)) +
 geom_boxplot(outliers = FALSE) +
  geom_jitter(width = jitter_width, size = point_size) + 
  theme_classic() + 
   theme(axis.text.x = element_text(angle = 90)) +
  labs(x = NULL, y = "Shannon-Wiener")

grid.arrange(p1, p2, p3, ncol = 3, bottom = "Country")
```


Trying the same but with genera.
```{r anovas_genera, eval=TRUE, fig.cap="Distribution of diversity indices of genera by country.", fig.height=10, fig.width=10}

plotlist <- list()
jitter_width <- 0.15
point_size <- 0.3

# Running model first for epifauna data, then infauna
for(set in c("Epifauna","Infauna")){
  
  dat_overview <- all_dat_env %>%
    filter(Set == set & !is.na(Genus) & Abundance >0) %>%
    filter(Abundance > 0)  %>%
    mutate(across(c(Country, Site, Month), as.factor)) %>%
    group_by(Country, Site, Year, Month, Genus)  %>%
    summarise(GenusAbundance = sum(Abundance)) %>% # Summing the total abundance per genus (if more than one)
    ungroup() %>%
    group_by(Country, Site, Year, Month)  %>%
    summarise(TotAbundance = sum(GenusAbundance),
      NumGenus = n_distinct(Genus),
      Shannon = diversity(GenusAbundance, index = "shannon")) %>%
    ungroup()

  print(paste0("Formulating ANOVAs for ", set))

  m_abun <- aov(log(TotAbundance+1) ~ Country + Year + Month + Country:Site, data = dat_overview)
  print(m_abun$call)
  print(summary(m_abun))

  m_genus <- aov(NumGenus ~ Country + Year + Month + Country:Site, data = dat_overview)
  print(m_genus$call)
  print(summary(m_genus))

  m_div <- aov(Shannon ~ Country + Year + Month + Country:Site, data = dat_overview)
  print(m_div$call)
  print(summary(m_div))

  p1 <- ggplot(dat_overview, aes(x = Country, y = log(TotAbundance+1))) +
    geom_boxplot(outliers = FALSE) +
    geom_jitter(width = jitter_width, size = point_size) + 
    theme_classic() + 
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x= NULL, y = "log(Abundance)")
  
  p2 <- ggplot(dat_overview, aes(x = Country, y = NumGenus)) +
    geom_boxplot(outliers = FALSE) +
    geom_jitter(width = jitter_width, size = point_size) + 
    theme_classic() + 
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x= NULL, y = "Number of genera")
  
  p3 <- ggplot(dat_overview, aes(x = Country, y = Shannon)) +
   geom_boxplot(outliers = FALSE) +
    geom_jitter(width = jitter_width, size = point_size) + 
    theme_classic() + 
     theme(axis.text.x = element_text(angle = 90)) +
    labs(x = NULL, y = "Shannon-Wiener")
  
  plotlist <- append(plotlist, list(p1, p2, p3))
  }

g <- grid.arrange(grobs=plotlist, ncol = 3, bottom = "Country")
ggsave(g, file = "../Graphs/boxplots_diversity_genera.jpg", width=190, height=190,units="mm")
rm(list=ls(pattern="m_"))
rm(plotlist, p1, p2, p3)
```

NMDS on genus level gives warning "Best solution was not repeated" - not reliable results?
```{r nmds_epifauna_genus, eval=TRUE, fig.cap="NMDS plot epifauna genera.", fig.height=10, fig.width=10}
salinity_levels <- c("Low (<10)","Medium (10-15)","High (>15)")

# Reshape data
dat_w <- all_dat_env %>% 
# Creating factor of salinity level
  filter(!is.na(salinity))  %>%
  mutate(salinity_factor = case_when(
      salinity < 10 ~ salinity_levels[1],
      salinity >= 10 & salinity < 15 ~ salinity_levels[2],
      salinity >= 15 ~ salinity_levels[3])) %>%
  mutate(salinity_factor = factor(salinity_factor, levels = salinity_levels, ordered = TRUE)) %>%
  #Remove missing data, select relevant data
  filter(!(Site == "Finsbo" & Year==2024)) %>% 
  filter(!is.na(Genus) & Set == "Epifauna" & Abundance>0) %>%
  dplyr::select(!c(Group, ScientificName)) %>%
  # Sum if multiple species within same genus
  group_by(Site,Country,Year,Month,Replicate,Genus,Lat,Long,swm,salinity_factor) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Convert to wide format
  pivot_wider(names_from = Genus, values_from = Abundance, values_fill = 0)  %>%   # Assuming missing value = 0
  arrange(Country) 

# Making community matrix
comm_mat <- as.matrix(dat_w  %>%
  dplyr::select(-(Site:salinity_factor)))

colnames(comm_mat) <- colnames(dat_w  %>% dplyr::select(-(Site:salinity_factor)))
rownames(comm_mat) <- paste(dat_w$Site,dat_w$Year,dat_w$Month,dat_w$Replicate, sep="_")

ord.fauna <- metaMDS(comm_mat)
#stressplot(ord.fauna)

# Plot the NMDS with site names as text labels
# Grouped by country
countries <- unique(dat_w$Country)
cols_countries <- rainbow(n = length(countries), alpha = 0.5)

jpeg("../Graphs/MNDS_epifauna_genera.jpg", width=190, height=190,units="mm",res=300)
plot(ord.fauna, type="n")
ordiellipse(ord.fauna, dat_w$Country, draw="polygon",label=FALSE, col = cols_countries, lty =0)
orditorp(ord.fauna, type="n" ,display="sites",cex=0.7)
legend("bottomright", legend = countries, col = cols, pch = 19, bty = "n")
ordiellipse(ord.fauna, dat_w$salinity_factor, draw="polygon",label=TRUE, lty =1:3,col="lightgrey",alpha=0.5)
dev.off()

```

NMDS for infauna data gives warning of possibly insufficient data ("stress is (nearly) zero") and not meaningful results. 
```{r nmds_infauna_genus, eval=TRUE, fig.cap="NMDS plot infauna genera.", fig.height=10, fig.width=10}
salinity_levels <- c("Low (<10)","Medium (10-15)","High (>15)")

# Reshape data
dat_w <- all_dat_env %>% 
# Creating factor of salinity level
  filter(!is.na(salinity))  %>%
  mutate(salinity_factor = case_when(
      salinity < 10 ~ salinity_levels[1],
      salinity >= 10 & salinity < 15 ~ salinity_levels[2],
      salinity >= 15 ~ salinity_levels[3])) %>%
  mutate(salinity_factor = factor(salinity_factor, levels = salinity_levels, ordered = TRUE)) %>%
  #Remove missing data, select relevant data
  filter(!(Site == "Finsbo" & Year==2024)) %>% 
  filter(!is.na(Genus) & Set == "Infauna" & Abundance>0) %>%
  dplyr::select(!c(Group, ScientificName)) %>%
  # Sum if multiple species within same genus
  group_by(Site,Country,Year,Month,Replicate,Genus,Lat,Long,swm,salinity_factor) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  ungroup()  %>%
  # Convert to wide format
  pivot_wider(names_from = Genus, values_from = Abundance, values_fill = 0)  %>%   # Assuming missing value = 0
  arrange(Country) 

# Making community matrix
comm_mat <- as.matrix(dat_w  %>%
  dplyr::select(-(Site:salinity_factor)))

colnames(comm_mat) <- colnames(dat_w  %>% dplyr::select(-(Site:salinity_factor)))
rownames(comm_mat) <- paste(dat_w$Site,dat_w$Year,dat_w$Month,dat_w$Replicate, sep="_")

ord.fauna <- metaMDS(comm_mat)
stressplot(ord.fauna)

# Plot the NMDS with site names as text labels
# Grouped by country
countries <- unique(dat_w$Country)
cols_countries <- rainbow(n = length(countries), alpha = 0.5)
cols_salinity <- rainbow(n = length(salinity_levels), alpha = 0.5)

plot(ord.fauna, type="n")
ordiellipse(ord.fauna, dat_w$Country, draw="polygon",label=FALSE, col = cols_countries, lty =0)
orditorp(ord.fauna, type="n" ,display="sites",cex=0.7)
legend("bottomright", legend = countries, col = cols, pch = 19, bty = "n")

ordiellipse(ord.fauna, dat_w$salinity_factor, draw="polygon",label=TRUE, lty =1:3)

```

```{r yeartrend_gam, eval=FALSE}
# Analyzing yearly trends using GAMs, not included
library(mgcv)

# Create dataframe for model predictions
range_years <-  dat_overview  %>%
  group_by(Country)  %>%
  summarise(start_year = min(Year), end_year = max(Year))

newdat <- data.frame(Country = rep(unique(dat_overview$Country), each = 56),
                     Year = seq(min(dat_overview$Year),
                                      max(dat_overview$Year))) %>%
  left_join(range_years)  %>%
  group_by(Country)  %>%
  filter(Year >= start_year & Year <= end_year) %>%
  ungroup() %>%
  dplyr::select(Country, Year)

# GAM models
gam_abun <- gam(log(TotAbundance+1) ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_abun)
# No significant effects when using raw data
# After log transformation, significant year effects for Norway

gam_groups <- gam(NumGroups  ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_groups)
# Borderline for Norway  

gam_div <- gam(Shannon  ~ s(Year, by = Country, k = 3) + Country , data = dat_overview)
summary(gam_div)
# No significant effects

# Model predictions

newdat <- newdat  %>%
  mutate(TotAbundance = predict.gam(gam_abun, newdata = newdat),
         NumGroups = predict.gam(gam_groups, newdata = newdat),
         Shannon = predict.gam(gam_div, newdata = newdat))

p1 <- ggplot(dat_overview, aes(x = Year, y = log(TotAbundance+1), color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = TotAbundance, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "top") +
  labs(x= NULL, y = "log(Abundance)")

p2 <- ggplot(dat_overview, aes(x = Year, y = NumGroups, color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = NumGroups, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "none") +
 labs(x= NULL, y = "Number of groups")

p3 <- ggplot(dat_overview, aes(x = Year, y = Shannon, color = Country)) +
  geom_point() +
  geom_line(data = newdat, aes(x = Year, y = Shannon, color = Country)) +
  theme_minimal() + 
  theme(legend.position = "none") +
  labs(x = NULL, y = "Shannon-Wiener")

grid.arrange(p1, p2, p3, ncol = 1, bottom = "Year")

```
