---
title: "SpeciesChecking"
output: html_document
date: "2025-04-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(plyr)
library(ggplot2)
load("output/infauna_epifauna_analysis_042025.RData")
```


# Epifauna

For the epifauna data we mainly focused on amphipods, isopods, gastropods, Mytilus, and epifaunal polychaetes. Orders that were not considered included , 


```{r}
checking_epi <- ddply(all_dat_env_t[all_dat_env_t$epifauna_or_infauna %in% c("Epifauna", "Combined"), ],
                      c("Site", "order", "class", "epifauna_or_infauna",
                        "family", "genus", "valid_name", "Country"), summarise,
                      N = sum(as.numeric(Abundance > 0)))
checking_epi <- checking_epi[checking_epi$N > 0, ]


##
species_dat <- ddply(all_dat_env_t[all_dat_env_t$epifauna_or_infauna %in% c("Epifauna", "Combined"), ],
                     c("Site", "order", "class",
                       "family", "genus", "Country"), summarise,
                     N = sum(as.numeric(Abundance > 0)))
species_dat <- species_dat[species_dat$N > 0, ]

unique(species_dat$order)
```

## Amphipoda

Germany, Norway and Sweden have amphipods that have not been identified to genus level. Based on further investigation, the following are consiered on family level: Ampithoidae, Aoridae, Calliopiidae, Caprellidae, Dexaminidae, Gammaridae, Hyalidae, Melitidae, Stenothoidae. Whereas Ischyroceridae clearly had different genuses in different countries.

```{r}
group <- "Amphipoda"
g <- ggplot(species_dat[species_dat$order %in% group, ], aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus") + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



```{r}

dat_t <- species_dat[species_dat$Country %in% c("Sweden W", "Germany", "Norway"), ]
hist_tab <- unique(dat_t[dat_t$order %in% group &
                           is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

table(dat_t[dat_t$family %in% hist_tab, "family"],
      dat_t[dat_t$family %in% hist_tab, "genus"],
      dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Genus") + ylab("Family") + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g

```

## Asteroidea


```{r}
group <- "Asteroidea"
g <- ggplot(species_dat[species_dat$class %in% group, ], aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus") + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```

## Ascidiacea

```{r}
group <- "Ascidiacea"
g <- ggplot(species_dat[species_dat$class %in% group, ], aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus") + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
# g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



## Isopoda

```{r}
group <- "Isopoda"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



```{r}

dat_t <- species_dat[species_dat$Country %in% c("Norway"), ]
hist_tab <- unique(dat_t[dat_t$order %in% group &
                           is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

table(dat_t[dat_t$family %in% hist_tab, "family"],
      dat_t[dat_t$family %in% hist_tab, "genus"],
      dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Genus") + ylab("Family") + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g

```

## Cephalaspidea

Cephalaspidea were considered on genus level, with the exception of Haminoeidae that was only identified to family.

```{r}
unique(species_dat$order)
group <- "Cephalaspidea"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



```{r}

dat_t <- species_dat[species_dat$Country %in% c("Norway", "Sweden W", "Germany"), ]
hist_tab <- unique(dat_t[dat_t$order %in% group &
                           is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

table(dat_t[dat_t$family %in% hist_tab, "family"],
      dat_t[dat_t$family %in% hist_tab, "genus"],
      dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Genus") + ylab("Family") + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g

```

## Decapoda

Palaemon considered on species level. Disregarded Paguridae, Inachidae, Hippolytidae and  Crangonidae.

```{r}
unique(species_dat$order)
group <- "Decapoda"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



```{r}

dat_t <- species_dat[species_dat$Country %in% c("Norway", "Sweden W", "Germany"), ]
hist_tab <- unique(dat_t[dat_t$order %in% group &
                           is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

table(dat_t[dat_t$family %in% hist_tab, "family"],
      dat_t[dat_t$family %in% hist_tab, "genus"],
      dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Genus") + ylab("Family") + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g

```

## Diptera

From Diptera only the Chironomidae family was included. 

```{r}
unique(species_dat$order)
group <- "Diptera"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```

## Littorinimorpha

Hydrobiidae with no information on genus were disregarded as the number of observations were very few and only in Germany. Rissoidae considered as family.

```{r}
unique(species_dat$order)
group <- "Littorinimorpha"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



```{r}

dat_t <- species_dat[species_dat$Country %in% c("Norway", "Sweden W", "Germany"), ]
hist_tab <- unique(dat_t[dat_t$order %in% group &
                           is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

table(dat_t[dat_t$family %in% hist_tab, "family"],
      dat_t[dat_t$family %in% hist_tab, "genus"],
      dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Genus") + ylab("Family")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g

```

## Mytilida

So few Mytilida that only considering genus

```{r}
unique(species_dat$order)
group <- "Mytilida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```

## Neogastropoda

```{r}
unique(species_dat$order)
group <- "Neogastropoda"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```


## Phyllodocida

Considered on family level

```{r}
unique(species_dat$order)
group <- "Phyllodocida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



## Cycloneritida


```{r}
unique(species_dat$order)
group <- "Cycloneritida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```


## Tanaidacea

Considering on genus level and disregarding family level observations (N = 1)

```{r}
unique(species_dat$order)
group <- "Tanaidacea"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



## Caenogastropoda incertae sedis

Considering Bittium on genus level and Turritellidae as family

```{r}
unique(species_dat$order)
group <- "Caenogastropoda incertae sedis"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```
## Mysida

Either species or order level data avilable; decided to go with order

```{r}
unique(species_dat$order)
group <- "Mysida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```
## Aplysiida

```{r}
unique(species_dat$order)
group <- "Aplysiida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```

## Nudibranchia

Considering as order


```{r}
unique(species_dat$order)
group <- "Nudibranchia"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



## Trochida

Considering genus and Skeneidae as family


```{r}
unique(species_dat$order)
group <- "Trochida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```


## Lymnaeidae


```{r}
unique(species_dat$order)
group <- "Lymnaeidae"

g <- ggplot(species_dat[species_dat$family %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```
## Other Gastropods

The following had no order and were considered either on family: Pyramidellidae, Limapontiidae, or genus level: Omalogyridae

## Genus NA


```{r}

dat_t <- species_dat
hist_tab <- unique(dat_t[is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

# table(dat_t[dat_t$family %in% hist_tab, "family"],
#       dat_t[dat_t$family %in% hist_tab, "genus"],
#       dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = order)) + geom_point()
g <- g + ggtitle("NA genus")
g <- g + xlab("Genus") + ylab("Family")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g

```


# Infauna

For the infauna data we mainly focused on infaunal polychaetes, oligochaetes, amphipods, and bivalvs. Orders that were not considered included , 

## Adapedonta

Only two genus

```{r}
checking <- ddply(all_dat_env_t[all_dat_env_t$epifauna_or_infauna %in% c("Infauna", "Combined"), ],
                  c("Site", "order", "class", "epifauna_or_infauna",
                    "family", "genus", "valid_name", "Country"), summarise,
                  N = sum(as.numeric(Abundance > 0)))
checking <- checking[checking$N > 0, ]


##
species_dat <- ddply(all_dat_env_t[all_dat_env_t$epifauna_or_infauna %in% c("Infauna", "Combined"), ],
                     c("Site", "order", "class",
                       "family", "genus", "Country"), summarise,
                     N = sum(as.numeric(Abundance > 0)))
species_dat <- species_dat[species_dat$N > 0, ]
```


```{r}
unique(species_dat$order)
group <- "Adapedonta"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```

## Amphipoda

Kept all genus for Corophiidae and for Norway there was only family available
Aoridae dropped as epifauna.
Caprellidae dropped as epifauna.
Dexamine dropped as epifauna.
Gammaridae dropped as epifauna.
Ischyroceridae dropped as epifauna.
Melitidae dropped as epifauna.



```{r}
unique(species_dat$order)
group <- "Amphipoda"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



```{r}

dat_t <- species_dat[species_dat$Country %in% c("Norway"), ]
hist_tab <- unique(dat_t[dat_t$order %in% group &
                           is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

table(dat_t[dat_t$family %in% hist_tab, "family"],
      dat_t[dat_t$family %in% hist_tab, "genus"],
      dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Genus") + ylab("Family")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g
```


## Cardiida

All Cardiidae considered only on family level as Parvicardium and Cerastoderma are difficult to distinguish when small.


```{r}
unique(species_dat$order)
group <- "Cardiida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



```{r}

dat_t <- species_dat[species_dat$Country %in% c("Norway", "Finland"), ]
hist_tab <- unique(dat_t[dat_t$order %in% group &
                           is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

table(dat_t[dat_t$family %in% hist_tab, "family"],
      dat_t[dat_t$family %in% hist_tab, "genus"],
      dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Genus") + ylab("Family")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g
```


## Myida

```{r}
unique(species_dat$order)
group <- "Myida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



## Phyllodocida

```{r}
unique(species_dat$order)
group <- "Phyllodocida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



```{r}

dat_t <- species_dat[species_dat$Country %in% c("Norway", "Finland"), ]
hist_tab <- unique(dat_t[dat_t$order %in% group &
                           is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

table(dat_t[dat_t$family %in% hist_tab, "family"],
      dat_t[dat_t$family %in% hist_tab, "genus"],
      dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Genus") + ylab("Family")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g
```

## Sabellida

```{r}
unique(species_dat$order)
group <- "Sabellida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



```{r}

dat_t <- species_dat[species_dat$Country %in% c("Norway", "Sweden W"), ]
hist_tab <- unique(dat_t[dat_t$order %in% group &
                           is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

table(dat_t[dat_t$family %in% hist_tab, "family"],
      dat_t[dat_t$family %in% hist_tab, "genus"],
      dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Genus") + ylab("Family")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g
```

## Spionida

Also keeping Spionida as family in addition to all genus levels.

```{r}
unique(species_dat$order)
group <- "Spionida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```



```{r}

dat_t <- species_dat[species_dat$Country %in% c("Norway", "Sweden W"), ]
hist_tab <- unique(dat_t[dat_t$order %in% group &
                           is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

table(dat_t[dat_t$family %in% hist_tab, "family"],
      dat_t[dat_t$family %in% hist_tab, "genus"],
      dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Genus") + ylab("Family")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g
```

## Galeommatida

```{r}
unique(species_dat$order)
group <- "Galeommatida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```


## Venerida

```{r}
unique(species_dat$order)
group <- "Venerida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```


## Eunicida

```{r}
unique(species_dat$order)
group <- "Eunicida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```


## Terebellida

```{r}
unique(species_dat$order)
group <- "Terebellida"

g <- ggplot(species_dat[species_dat$order %in% group, ],
            aes(Site, genus, size = N, col = family)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Genus")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```
## Genus NA


```{r}

dat_t <- species_dat
hist_tab <- unique(dat_t[is.na(dat_t$genus) &
                           !is.na(dat_t$family), ]$family)

# table(dat_t[dat_t$family %in% hist_tab, "family"],
#       dat_t[dat_t$family %in% hist_tab, "genus"],
#       dat_t[dat_t$family %in% hist_tab, "Country"])

dat_t[is.na(dat_t$genus), "genus"] <- "NA"

g <- ggplot(dat_t[dat_t$family %in% hist_tab, ], aes(genus, family, col = order)) + geom_point()
g <- g + ggtitle("NA genus")
g <- g + xlab("Genus") + ylab("Family")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g
```
## Clitellata

```{r}

group <- "Clitellata"

g <- ggplot(species_dat[species_dat$class %in% group, ],
            aes(Site, order, size = N, col = class)) + geom_point()
g <- g + ggtitle(group)
g <- g + xlab("Site") + ylab("Order")
g <- g + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
g <- g + facet_wrap(~Country, nrow = 1, scales = "free_x")
g


```
